---
title: "Conditional Simulation in spmodel"
author: "Michael Dumelle, Matt Higham, and Jay M. Ver Hoef"
bibliography: '`r system.file("references.bib", package="spmodel")`'
output:
  html_document:
editor_options:
  chunk_output_type: console
---

```{r setup, include = FALSE}
# # jss style
# knitr::opts_chunk$set(prompt=TRUE, echo = TRUE, highlight = FALSE, continue = " + ", comment = "")
# options(replace.assign=TRUE, width=90, prompt="R> ")

# rmd style
knitr::opts_chunk$set(collapse = FALSE, comment = "#>", warning = FALSE, message = FALSE)

# loading
library(spmodel)
library(sf)
library(ggplot2)
```

# Introduction {#sec:introduction}

`spmodel` is an **R** package used to fit, summarize, and
predict for a variety of spatial statistical models. This vignette explores `conditional()`, which many plausible random samples of model parameters and `newdata`, an approach known as conditional simulation. Before proceeding, we
load `spmodel`, `sf`, and `ggplot2` by running
```{r, eval = FALSE}
library(spmodel)
library(sf)
library(ggplot2)
```

If using `spmodel` in a formal publication or report, please cite it. Citing `spmodel` lets us devote more resources to it in the future. We view the `spmodel` citation by running

```{r}
citation(package = "spmodel")
```

Before proceeding, we set a reproducible seed:
```{r}
set.seed(0)
```

# Conditional Simulation in `spmodel`

`predict()` returns a single best prediction (and its standard error) at each new location, one location at a time. However, `predict()` does not tell us how predictions at different locations may relate to one another. Suppose we instead want to know something about the entire predicted surface at once (e.g., quantiles, extreme values, threshold exceedances, etc.). Answering these types of questions requires jointly simulating the predicted surface, not just each location's pointwise mean and standard error. `conditional()` provides exactly this: repeated draws from the joint distribution of the response (or, for `spglm()`/`spgautor()` models, the latent process) at a set of new locations, conditional on the observed data. Though typically the focus on conditional simulation is on predictions at new locations, `conditional()` also provides simulated values of covariance and fixed effect model parameters. For technical details regarding conditional simulation, see @zimmerman2024spatial and the Techincal Details vignette.

The `sulfate` data is an `sf` object that contains sulfate measurements in the conterminous United States (CONUS). We first visualize the distribution of the sulfate data by running
```{r}
ggplot(sulfate, aes(color = sulfate)) +
  geom_sf(size = 2.5) +
  scale_color_viridis_c() +
  theme_gray(base_size = 18)
```

We fit an intercept-only spatial linear model with an exponential spatial covariance function by running
```{r}
sulfmod <- splm(sulfate ~ 1, data = sulfate, spcov_type = "exponential")
```

We draw 1,000 conditional simulations at every location in `sulfate_preds` by running
```{r}
sulf_cond <- conditional(sulfmod, newdata = sulfate_preds)
dim(sulf_cond)
```

`sulf_cond` is a matrix with one row per row of `sulfate_preds` and one column per sample (the default number of samples is 1,000, controlled by the `samples` argument). Each row is 1,000 draws from the predictive distribution of sulfate at that location, and because the draws are joint (all locations are simulated together for each sample/column), the spatial dependence among locations is preserved within each column.

The 20th row of `sulfate_preds` looks like:

```{r}
sulfate_preds[20, ]
```

At the 20th row of `sulfate_preds`, the mean and standard deviation from `predict()` closely match the same statistics calculated from the conditional simulations:
```{r}
predict(sulfmod, sulfate_preds[20, ], se.fit = TRUE)
```

```{r}
fit_cond <- mean(sulf_cond[20, ])
se_cond <- sd(sulf_cond[20, ])
c(fit_cond = fit_cond, se.fit_cond = se_cond)
```

We can visualize the distribution of conditional simulations at this location:
```{r}
ggplot(data.frame(sulfate20 = sulf_cond[20, ]), aes(x = sulfate20)) +
  geom_histogram(bins = 30) +
  theme_bw(base_size = 14)
```


## Using the Conditional Simulations

Using conditional simulation, we can compute quantities of this predictive distribution, like quantiles, that are not able to be computed via `predict()`:

```{r}
q_10 <- as.numeric(quantile(sulf_cond[20, ], probs = 0.1))
q_median <- median(sulf_cond[20, ])
q_95 <- as.numeric(quantile(sulf_cond[20, ], probs = 0.95))
c(q_10 = q_10, q_median = q_median, q_95 = q_95)
```

Now suppose we want to know the proportion of `sulfate_preds` locations with sulfate concentration above 5. Again, this nonlinear function of the predicted surface cannot be quantified by `predict()`'s pointwise estimates and standard errors alone. Using the conditional simulations, we compute this proportion once per sample (i.e., for each column of `sulf_cond`) to obtain a distribution of the proportion itself:
```{r}
prop_exceed <- apply(sulf_cond, 2, function(x) mean(x > 5))
```

Then we can return the mean and standard error of this proportion:
```{r}
fit_cond <- mean(prop_exceed)
se_cond <- sd(prop_exceed)
c(fit_cond = fit_cond, se.fit_cond = se_cond)
```

We can also compute a quantile-based 95% confidence interval for the proportion:
```{r}
ci_bounds <- as.numeric(quantile(prop_exceed, probs = c(0.025, 0.975)))
c(lwr = ci_bounds[1], upr = ci_bounds[2])
```


## Propagating Covariance Parameter Uncertainty

By default, every draw in `sulf_cond` shares the same fitted covariance parameters -- only the fixed effects and the spatial field are resimulated. For `splm()` model objects fit with `ddf = "satterthwaite"` (the default when the sample size is 500 or fewer, as it is here), setting `simulate_covparams = TRUE` additionally resamples the covariance parameters themselves for each draw, propagating their own estimation uncertainty into the result. This is considerably more expensive than the default (a fresh factorization is required for every sample), so it is generally reserved for small-to-moderate sample sizes and a reduced `samples`.

We repeat the earlier `sulfmod` simulation with `simulate_covparams = TRUE`:
```{r}
sulf_cond_cp <- conditional(sulfmod, newdata = sulfate_preds, samples = 200, simulate_covparams = TRUE)
```

Because this extra layer of uncertainty is additive, the standard deviation of the draws at a given location is larger than under the default (fixed-covariance-parameter) simulation:
```{r}
c(default = sd(sulf_cond[20, ]), simulate_covparams = sd(sulf_cond_cp[20, ]))
```

## Additional Output

By default, `conditional()` returns simulated values for `newdata` (`output = "newdata"`, the default). Two other outputs are available and can be combined with `"newdata"` (or with each other) by supplying a character vector to `output`:

* `"beta"`: The simulated fixed effect coefficients used to generate each sample.
* `"object"`: The observed data, replicated once per sample (useful for row-binding observed and predicted values together).

For example, we can get simulated `newdata` values and their corresponding simulated fixed effects together by running
```{r}
cond_list <- conditional(sulfmod, newdata = sulfate_preds, samples = 200, simulate_covparams = TRUE, output = c("newdata", "beta", "cov"))
names(cond_list)
dim(cond_list$beta)
dim(cond_list$cov)
```

We can visualize, for example, the conditional simulations of the spatial range parameter:
```{r}
ggplot(data.frame(range = cond_list$cov[3, ]), aes(x = range)) +
  geom_histogram(bins = 30) +
  theme_bw(base_size = 14)
```

# Spatial Generalized Linear Models

`conditional()` also works for `spglm()` and `spgautor()` model objects. Because these models describe a latent process $\mathbf{w}$ on the link scale rather than the response directly, the `type` argument controls the scale of the simulated values: `type = "link"` (the default) returns simulations of $\mathbf{w}$, `type = "response"` returns simulations of the mean on the response (inverse link) scale, and `type = "new"` additionally simulates a new observation from the response distribution (e.g., a new Poisson count) with that mean.

We fit a spatial logistic regression model of moose presence and simulate new observations at the locations in `moose_preds`:
```{r}
binmod <- spglm(presence ~ elev, family = "binomial", data = moose, spcov_type = "exponential")
moose_cond <- conditional(binmod, newdata = moose_preds, type = "new")
```

Each row of `moose_cond` now contains 1,000 simulated presences for that location rather than an underlying proportion, reflecting both the uncertainty in the underlying probability surface and the additional sampling variability for an individual observation from the binomial distribution.

We can visualize the distribution of conditional simulations at the first location:

```{r}
ggplot(data.frame(presence = moose_cond[1, ]), aes(x = presence)) +
  geom_histogram(bins = 30) +
  theme_bw(base_size = 14)
```

# Big Data

Both fitting a model to many observed locations and conditionally simulating at many new locations can become computationally expensive as sample sizes grow. We simulate 6,000 observed locations and fit a spatial linear model, requesting `splm()`'s own big data approximation via `local = TRUE`:
```{r}
n <- 6000
big_dat <- data.frame(xc = runif(n), yc = runif(n))
big_dat$z <- sprnorm(
  spcov_params("exponential", de = 1, ie = 0.1, range = 0.2),
  data = big_dat, xcoord = xc, ycoord = yc
)
bigmod <- splm(z ~ 1, data = big_dat, spcov_type = "exponential", xcoord = xc, ycoord = yc, local = TRUE)
```

`conditional()` accommodates big data the same way `sprnorm()` does (see the "Simulating Large Spatial Data" vignette): via its own `local` argument, offering a low-rank and a Vecchia-style sequential approximation. We conditionally simulate 500 new locations using the low-rank approximation:
```{r}
big_newdata <- data.frame(xc = runif(500), yc = runif(500))
cond_lowrank <- conditional(bigmod, newdata = big_newdata, local = list(approximation = "low-rank"))
dim(cond_lowrank)
```

and, for comparison, using the Vecchia approximation:
```{r}
cond_vecchia <- conditional(bigmod, newdata = big_newdata, local = list(approximation = "vecchia"))
dim(cond_vecchia)
```

As with `predict()` and `sprnorm()`, `local = TRUE` is shorthand for the default implementation of the big data approach, which currently uses the low-rank approximation.

# References {.unnumbered}

<div id="refs"></div>
