---
title: "The Spatial Decorrelation Transformation for Machine and Deep Learning Models"
author: "Michael Dumelle, Matt Heaton, and Ryan A. Hill"
bibliography: '`r system.file("references.bib", package="spmodel")`'
output:
  html_document:
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(spmodel)
library(sf)
library(ggplot2)
library(ranger)
```

# Introduction {#sec:introduction}

`spmodel` is a **R** package used to fit, summarize, and
predict for a variety of spatial statistical models. This vignette
explores tools for a spatial decorrelation transformation that can be applied as a preprocessing step for machine and deep learning models like random forests. The spatial decorrelation works by first decorrelating the data by removing spatial covariance (i.e., correlation, dependence). These decorrelated data are used to fit a machine learning model and use it to make predictions of new observations. These predictions are then adjusted by recorrelating the data by adding back in the original spatial covariance. This approach is flexible and can accommodate any machine learning model, as the decorrelation and recorrelation steps happen separately from the machine learning modeling. The spatial decorrelation transformation can significantly improve a machine learning model's inferential and predictive capacity by formally incorporating spatial covariance into the model building process [@heaton2025adjusting].

Before proceeding, we load  `spmodel`, `sf`, `ggplot2`, and `ranger`:
```{r, eval = FALSE}
library(spmodel)
library(sf)
library(ggplot2)
library(ranger)
```

If using `spmodel` in a formal publication or report, please cite it.
Citing `spmodel` lets us devote more resources to the package in the
future. We view the `spmodel` citation:

```{r}
citation(package = "spmodel")
```

# The `lake` Data

In this vignette, we use the point-referenced `lake` data, an `sf` object that contains data on climate data for lakes in the southwestern United States. We view the first few rows of `lake`:
```{r}
lake
```
We can learn more about `lake` by running `help("lake", "spmodel")`.

Our goal is to use a machine learning model (here, a random forest) to study the impact of the explanatory variable(s) (here, temperature) on the response variable (here, log conductivity) while accounting for spatial covariance among nearby observations. Before fitting any models, we will visualize some relationships among variables. First, we visualize the spatial distribution of log conductivity (`log_cond`) in `lake` by running
```{r lake, fig.cap="Distribution of log conductivity in the data.", out.width = "65%", fig.align="center"}
ggplot(lake, aes(color = log_cond)) +
  geom_sf() +
  scale_color_viridis_c() +
  theme_gray(base_size = 14)
```

`log_cond` values are spatially patterned, whereby groups of high and low values of log conductivity occur together. Second, we visualize the relationship between log conductivity and temperature:
```{r}
ggplot(lake, aes(x = temp, y = log_cond)) +
  geom_point() +
  theme_gray(base_size = 14)
```

There appears to be a linearly increasing relationship between log conductivity and temperature. 

Ultimately, we will use our machine learning model to predict log conductivity at new lakes in `lake_preds`. We can view the first few rows of `lake_preds`:
```{r}
lake_preds
```

We can also view the `lake_preds` locations (large triangles) alongside hte `lake` locations (small circles):
```{r}
ggplot() +
  geom_sf(data = lake) +
  geom_sf(data = lake_preds, pch = 17, size = 5) +
  theme_gray(base_size = 14)
```


Fitting machine learning models often involves some random components (i.e., randomly subsampling variables), so we set a reproducible seed to reproduce results locally:
```{r}
set.seed(0)
```

# The Spatial Decorrelation Transformation and `decorrelate()`

## The `decorrelate()` Function

The `decorrelate()` function applies the spatial decorrelation transformation directly to a (training) data set. `decorrelate()` generally requires the following three arguments:

* `formula`: a formula that describes the relationship between the response variable ($\mathbf{y}$) and explanatory variables ($\mathbf{X}$)
* `data`: a `data.frame` or `sf` object that contains the response variable, explanatory variables, and spatial information
* `spcov_type`: the spatial covariance type (`"exponential"`, `"matern"`, `"car"`, etc)

```{r}
decorr_mod <- decorrelate(
  formula = log_cond ~ temp,
  data = lake,
  spcov_type = "exponential"
)
```

If `data` is an `sf` [@pebesma2018sf] object, spatial information is stored in the object's geometry. If `data` is a `data.frame`, then the x-coordinates and y-coordinates must also be provided via the `xcoord` and `ycoord` arguments to `decorrelate()`.

The `decorrelate()` estimates spatial decorrelation transformation parameters using a grid search approach [@heaton2025adjusting]. Observations in `data` are randomly assigned to a training or test set. By default, 75% of the observations are assigned to the training set and the remaining 25% to the test set. For each set of parameters in the grid search, the following steps are applied:

1. The spatial decorrelation transformation is applied to the training set 
2. A machine learning algorithm (by default, a random forest) is fit to the transformed training data set
3. The spatial decorrelation transformation is applied to the test set
4. The fitted machine learning algorithm is used to predict the response variable for the transformed test set
5. The predictions are recorrelated back to the original (i.e., untransformed) scale

The grid element whose parameters yield the lowest root-mean-squared-prediction error (RMSPE) in the test set are considered the "estimated" parameters. Using those estimated parameters, the spatial decorrelation transformation is applied to the entire data set (all observations in `data`) and stored in the `decorrelate_data` element of `decorrelate()`. Several other useful pieces of information are stored in the `decorrelate()` output, such as the evaluated grid and machine learning algorithm used.


## Prediction Using `predict()`



## Comparing to a Standard (Independent) Random Forest

```{r}
decorr_mod <- decorrelate(
  formula = log_cond ~ temp,
  data = lake,
  spcov_type = "exponential",
  training = list(training_index = seq(1, 60))
)
decorr_mod$training_index
decorr_mod$test_index
```

```{r}
rf_mod <- decorrelate(
  formula = log_cond ~ temp,
  data = lake,
  spcov_type = "none"
)

rf_mod$test_rmspe
decorr_mod$test_rmspe
```

## Additional Features

```{r}
decorr_mod <- decorrelate(
  formula = log_cond ~ temp,
  data = lake,
  spcov_type = "exponential",
  algorithm = "xgboost"
)
```


```{r}
decorr_mod <- decorrelate(
  formula = log_cond ~ temp,
  data = lake,
  spcov_type = "exponential",
  random = ~ origin
)
```


```{r}
decorr_mod <- decorrelate(
  formula = log_cond ~ temp,
  data = lake,
  spcov_type = "exponential",
  local = TRUE
)
```

# Building Your Own Spatial Decorrelation Transformation

## Step 1: Split the Obsered Data Into Training and Test Sets

Evaluating the effectiveness of a machine learning model is a multi-step process. First, we split the `lake` data into separate training and test sets. The goal is to fit a model to the training set and use it to predict the response variable for each datum in the test set. Because the test set contains the measured value of the response variable, the accuracy of the predictions generated using the test set can be directly evaluated using a metric like root-mean-squared-prediction error or predictive $R^2$ (i.e., the squared correlation between observations and predictions). 

We proceed by randomly assigning 75% of the `lake` data to the training set and the remaining 25% to the test set:
```{r}
# number of observations in the lake data
n <- NROW(lake)
# creating an index that corresponds with the row position of each datum
index <- seq(from = 1, to = n)
# setting the size of the training data
n_train <- floor(0.75 * n)
# randomly generating the indices in lake that are assigned to the training data
train_index <- sample(index, size = n_train)
# creating the training data
lake_train <- lake[train_index, ]
# creating the test data
lake_test <- lake[-train_index, ]
# number of rows in each?
NROW(lake_train)
NROW(lake_test)
```

# The Spatial Decorrelation Transformation2

The spatial decorrelation works by first decorrelating the data by removing spatial covariance (i.e., correlation, dependence). These decorrelated data are used to fit a machine learning model and use it to make predictions of new observations. These predictions are then adjusted by recorrelating the data by adding back in the original spatial covariance. This approach is flexible and can accommodate any machine learning model, as the decorrelation and recorrelation steps happen separately from the machine learning modeling. We first explore how the spatial decorrelation transformation works for a fixed set of spatial covariance parameters. We then show how to relax this assumption and actually estimate unknown spatial covariance parameters using a grid search.

## Assuming Parameters are Known

Suppose we had some prior information which led us to assume the response variable's true spatial covariance was an exponential type with a spatially dependent variance of 1.2, a range of 200,000 meters (200km), and an independent variance of 0.2. Let's store this as an `spcov_params` object:
```{r}
val <- spcov_params(
  spcov_type = "exponential",
  de = 1.2,
  range = 2e5,
  ie = 0.2
)
```

The data are spatially decorrelated using `decorrelate_data()`, which requires the following three arguments:

* `formula`: a formula that describes the relationship between the response variable and the explanatory variables
* `data`: a `data.frame` or `sf` object that contains the response variable, explanatory variables, and spatial information
* `spcov_params`: the spatial covariance parameters as an `spcov_params()` object

If `data` is an `sf` [@pebesma2018sf] object, spatial information is stored in the object's geometry. If `data` is a `data.frame`, then the x-coordinates and y-coordinates must also be provided via the `xcoord` and `ycoord` arguments to `decorrelate_data()`.

The relationship between log conductivity (the response variable) and temperature (the explanatory variable) is captured using **R**'s formula object passed to `decorrelate_data()`:
```{r}
# create a formula argument
form <- log_cond ~ temp
# decorrelate
tlake_train <- decorrelate_data(
  formula = form,
  data = lake_train,
  spcov_params = val
)
```

At this stage, we can fit a machine learning model to the decorrelated data. Here, we will fit a random forest using the `ranger` **R** package [@wright2015ranger]. We will pass to `ranger` the decorrelated explanatory variable matrix to the `x` argument and the decorrelated response variable to the `y` argument:
```{r}
rfmod <- ranger(x = tlake_train$tX, y = tlake_train$ty)
```

We could inspect the random forest using any of the standard functions that interact with `ranger` objects, such as `print()` or variable importance via `importance()`:
```{r}
print(rfmod)
rfmod_imp <- update(rfmod, importance = "impurity")
ranger::importance(rfmod_imp)
```

Before using `rfmod` to make predictions for the test data, we must decorrelate the explanatory variable in the test data using `decorrelate_newdata()` so they are on the same scale as the decorrelated training data:
```{r}
tlake_test <- decorrelate_newdata(tlake_train, newdata = lake_test)
```

Then, we can use `predict()` to make predictions for the decorrelated test data and recorrelate them using `recorrelate_newdata()`:
```{r}
rf_tpreds <- predict(rfmod, data = tlake_test$tX_newdata)$predictions
sp_decorr_preds <- recorrelate_newdata(tlake_test, rf_tpreds)
```

We can evaluate the accuracy of our predictions using metrics like root-mean-squared-prediction error (RMSPE) and predictive $R^2$:
```{r}
get_rmspe_cor2 <- function(y, preds) {
  errors <- y - preds
  c("rmspe" = sqrt(mean(errors^2)), "cor2" = cor(y, preds)^2)
}
decorr_metrics <- get_rmspe_cor2(lake_test$log_cond, sp_decorr_preds)
print(decorr_metrics)
```

There is a noticeable positive correlation between the predictions of log conductivity and the observations themselves:
```{r}
lake_test$decorr_preds <- sp_decorr_preds
ggplot(lake_test, aes(x = decorr_preds, y = log_cond)) +
  geom_abline(intercept = 0, slope = 1, color = "orange", linewidth = 2) +
  geom_point(size = 2) +
  theme_gray(base_size = 14)
```

## Estimating Parameters Using a Grid Search

In practice, it is rare that the decorrelations parameters are actually known. Hence, we must generally have methods to estimate these parameters. One method for estimating these parameters using a grid search. In a grid search, a test set metric (here, RMSPE) is evaluated using several different parameter combinations. The parameter combination yielding the best test set metric is taken as the estimtaed parameter set. Grid searches can also be used to find initial values in some sort of optimization algorithm.

Parameter combinations can be created and evaluated using custom code, though the `decorrelate_grid()` function provides a grid which explores differet, plausible parameter combinations:
```{r}
grid <- decorrelate_grid(
  formula = form,
  data = lake_train,
  spcov_type = "exponential"
)
grid
```

The magnitudes of the `de` and `ie` variance parameters are based on the sample variance from an independent linear model fit using `lm()`, and the magnitude of the `range` parameters are based on whether or not spatial covariance decays to approximately zero in the spatial domain our outside the spatial domain. 

We can then apply the spatial decorrelation transformation to each element of the grid and compute the test set metric, RMSPE. We will simultaneously compute predictive $R^2$:
```{r}
y <- lake_test$log_cond
n_grid <- NROW(grid)
grid$rmspe <- rep(0, n_grid)
grid$cor2 <- rep(0, n_grid)
for (i in seq(1, n_grid)) {
  grid_row <- grid[i, ]
  spcov_params_row <- spcov_params(
    spcov_type = grid_row[["spcov_type"]],
    de = grid_row[["de"]],
    ie = grid_row[["ie"]],
    range = grid_row[["range"]]
  )
  tlake_train <- decorrelate_data(form, lake_train, spcov_params_row)
  rfmod <- ranger(x = tlake_train$tX, y = tlake_train$ty)
  tlake_test <- decorrelate_newdata(tlake_train, newdata = lake_test)
  rf_tpreds <- predict(rfmod, data = tlake_test$tX_newdata)$predictions
  sp_decorr_preds <- recorrelate_newdata(tlake_test, rf_tpreds)
  errors <- y - sp_decorr_preds
  grid$rmspe[i] <-  sqrt(mean(errors^2))
  grid$cor2[i] <- cor(sp_decorr_preds, y)^2
}
grid
```

We estimate the parameters via the combination that yields the smallest RMSPE:
```{r}
best_rmspe <- which.min(grid$rmspe)
best_rmspe
best_rmspe_row <- grid[best_rmspe, ]
best_rmspe_row
```

and store these parameters for future use with `decorrelate_data()`, `decorrelate_newdata()`, and `recorrelate_newdata()` to make predictions at the lakes in `lake_preds`:
```{r}
spcov_params_use <- spcov_params(
  spcov_type = best_rmspe_row[["spcov_type"]],
  de = best_rmspe_row[["de"]], 
  ie = best_rmspe_row[["ie"]],
  range = best_rmspe_row[["range"]]
)
spcov_params_use
```

## Additional Grid Search Features

Anisotropy and random effects can also be applied to the grid creation via the `anisotropy` and `random` arguments, whose values can be provided to `decorrelate_data()` using the `spcov_params` and `randcov_params` arguments:
```{r}
grid <- decorrelate_grid(
  formula = form,
  data = lake_train,
  spcov_type = "exponential",
  anisotropy = TRUE,
  random = ~ year
)
NROW(grid)
head(grid, n = 5)
```

## Comparing to Other Approaches

We can compare test set RMSPE from the spatial decorrelation transformation to test set RMSPE from other common approaches like an independent (standard) random forest and a spatial residual random forest.

First we fit an independent random forest. We must remove the spatial geometry structure from `lake_train` and `lake_test`, as `ranger()` does not accept `sf` objects:
```{r}
lake_train_df <- st_drop_geometry(lake_train)
lake_test_df <- st_drop_geometry(lake_test)
```

The we fit a random forest model to the training data, predict log conductivity in the test data, and compute RMSPE and predictive $R^2$:
```{r}
indrfmod <- ranger(form, data = lake_train_df)
indrf_preds <- predict(indrfmod, data = lake_test_df)$predictions
indrf_metrics <- get_rmspe_cor2(lake_test$log_cond, indrf_preds)
```

Similarly, we fit a spatial residual model [@fox2020comparing] to the training data, predict log conductivity in the test data, and compute RMSPE and predictive $R^2$:
```{r}
splmRFmod <- splmRF(form, data = lake_train, spcov_type = "exponential")
splmRF_preds <- predict(splmRFmod, newdata = lake_test)
splmRF_metrics <- get_rmspe_cor2(lake_test$log_cond, splmRF_preds)
```

Comparing the three approaches reveals that the spatial decorrelation transformation has the lowest (best) RMSPE and highest (best) predictive $R^2$, followed by, unsurprisingly, the spatial residual random forest and finally by the independent (standard) random forest:
```{r}
dat <- data.frame(
  method = c("decorr", "indrf", "splmRF"),
  rmspe = c(best_rmspe_row[["rmspe"]], indrf_metrics[["rmspe"]], splmRF_metrics[["rmspe"]]),
  cor2 = c(best_rmspe_row[["cor2"]], indrf_metrics[["cor2"]], splmRF_metrics[["cor2"]])
)
dat[order(dat$rmspe), ]
```

## Predicting at Unobserved Lakes 

A primary goal is to use the best machine learning model to make predictions of log conductivity at the lakes in `lake_preds`. We can visualize the lakes in `lake` alongside those in `lake_preds`:
```{r}
ggplot() +
  geom_sf(data = lake, aes(color = log_cond)) +
  geom_sf(data = lake_preds, shape = 17, size = 5) +
  scale_color_viridis_c() +
  theme_bw(base_size = 14)
```

Using the estimate decorrleation parameter, we decorrelate the `lake` data (according to `formula`):
```{r}
tlake <- decorrelate_data(form, lake, spcov_params = spcov_params_use)
```

Then, we fit a random forest model to the decorrelated data and make log conductivity predictions (on the decorrelated scale):
```{r}
rfmod <- ranger(x = tlake$tX, y = tlake$ty)
tlake_preds <- decorrelate_newdata(tlake, newdata = lake_preds)
rf_tpreds <- predict(rfmod, data = tlake_preds$tX_newdata)$predictions
```

Finally, we recorrelate the log conductivity predictions back to the original scale and visualize them alongside the log conductivity values in `lake`:
```{r}
lake_preds$preds <- recorrelate_newdata(tlake_preds, rf_tpreds)
ggplot() +
  geom_sf(data = lake, aes(color = log_cond)) +
  geom_sf(data = lake_preds, aes(color = preds), shape = 17, size = 5) +
  scale_color_viridis_c() +
  theme_bw(base_size = 14)
```

The log conductivity predictions closely resemble patterns from the observed lakes.

# References {.unnumbered}

<div id="refs"></div>
