---
title: "The Spatial Decorrelation Transformation for Machine Learning Models"
author: "Michael Dumelle, Matt Higham, and Jay M. Ver Hoef"
bibliography: '`r system.file("references.bib", package="spmodel")`'
output:
  html_document:
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(spmodel)
library(sf)
library(ggplot2)
library(ranger)
```

# Introduction {#sec:introduction}

`spmodel` is a **R** package used to fit, summarize, and
predict for a variety of spatial statistical models. This vignette
explores tools for a spatial decorrelation transformation that can be applied as a preprocessing step for machine learning models like random forests. The spatial decorrelation transformation can significantly improve a machine learning model's inferential and predictive capacity [@heaton2024adjusting].

Before proceeding, we load  `spmodel`, `sf`, `ggplot2`, and `ranger` by running
```{r, eval = FALSE}
library(spmodel)
library(sf)
library(ggplot2)
library(ranger)
```

If using `spmodel` in a formal publication or report, please cite it.
Citing `spmodel` lets us devote more resources to the package in the
future. We view the `spmodel` citation by running

```{r}
citation(package = "spmodel")
```

# The `moss` Data

In this vignette, we use the point-referenced `moss` data, an `sf` object that contains data on heavy metals in mosses near a mining road in Alaska. We view the first few rows of `moss` by running
```{r}
moss
```
We can learn more about `moss` by running `help("moss", "spmodel")`.

Our goal is to use a machine learning model (here, a random forest) to study the impact of the explanatory variables (log distance from the mining road and whether the datum is north or south of the mining road) on the response variable (log zinc concentration) while accounting for spatial covariance among nearby observations. Before fitting any models, we will visualize some relationships among these variables. First, we visualize the spatial distribution of log zinc concentration (`log_Zn`) in `moss` by running
```{r log_zn, fig.cap="Distribution of log zinc concentration in the moss data.", out.width = "65%", fig.align="center"}
ggplot(moss, aes(color = log_Zn)) +
  geom_sf() +
  scale_color_viridis_c() +
  theme_gray(base_size = 14)
```

`log_Zn` values are highest near the mining road in the center of the spatial domain. Second, we visualize the relationship between log zinc concentration (`log_Zn`) and both the (logarithm of) distance from the road (`log_dist2road`) and the side of the road (`sideroad`) by running
```{r}
ggplot(moss, aes(x = log_dist2road, y = log_Zn, color = sideroad, shape = sideroad)) +
  geom_point() +
  theme_gray(base_size = 14)
```

There appears to be a linearly decreasing relationship between the `log_dist2road` and `log_Zn`.

Finally, we set a seed so that vignette output can be reproduced locally:
```{r}
set.seed(0)
```


# Creating Training and Test Data

Evaluating the effectiveness of a machine learning model is a multi-step process. First, we split our data up into separate training and test sets. The training data are used to build a model and use it to predict the response variable for each datum in the test data. These predictions are compared to the true, realized value of the response variable using a metric like root-mean-squared-prediction error or predictive R-squared. 

We use a random 75/25 split for our training/test data, meaning that 75% of the observations are randomly assigned to the training data and 25% to the test data:
```{r}
# number of observations in the moss data
n <- NROW(moss)
# creating an index that corresponds with the row position of each datum
index <- seq(from = 1, to = n)
# setting the size of the training data
n_train <- floor(0.75 * n)
# randomly generating the indices in moss that are assigned to the training data
train_index <- sample(index, size = n_train)
# creating the training data
moss_train <- moss[train_index, ]
# creating the test data
moss_test <- moss[-train_index, ]
```

# The Spatial Decorrelation Transformation

The spatial decorrelation transformation removes spatial covariance (i.e., correlation) induced by some spatial covariance function. Suppose that we knew the response variable's true spatial covariance was exponential with a spatially dependent variance of 1.2, a range of 50,000 meters, and an independent variance of 0.2. Let's store this as an `spcov_params` object:
```{r}
val <- spcov_params(
  spcov_type = "exponential",
  de = 1.2,
  range = 5e4,
  ie = 0.2
)
```


The spatial decorrelation transformation is applied using `decorrelate_data()`, which relies on at least the following three arguments:

* `formula`: a formula that describes the relationship between the response variable and the explanatory variables
* `data`: a `data.frame` or `sf` object that contains the response variable, explanatory variables, and spatial information
* `spcov_params`: the spatial covariance parameters as an `spcov_params()` object

If `data` is an `sf` [@pebesma2018sf] object, spatial information is stored in the object's geometry. If `data` is a `data.frame`, then the x-coordinates and y-coordinates must be provided via the `xcoord` and `ycoord` arguments.

We apply the spatial decorrelation transformation to the training data according to `formula` and `spcov_params` by running
```{r}
form <- log_Zn ~ log_dist2road + sideroad
tmoss_train <- decorrelate_data(
  formula = form,
  data = moss_train,
  spcov_params = val
)
```

Now we can fit a machine learning model, here a random forest using the ranger **R** package [@wright2015ranger], to the transformed response variable (`tmoss_train$ty`) using the transformed explanatory variables (stored as a matrix in `tmoss_train$tX`):
```{r}
rfmod <- ranger(x = tmoss_train$tX, y = tmoss_train$ty)
```

Before using `rfmod` to make predictions for the test data, we must transform the explanatory variables in the test data using `decorrelate_newdata()` so they are on the same scale as the transformed training data:
```{r}
tmoss_test <- decorrelate_newdata(tmoss_train, newdata = moss_test)
rf_tpreds <- predict(rfmod, data = tmoss_test$tX_newdata)$predictions
```

Then, we can take these test data predictions and recorrelate them with spatial covariance using `recorrelate()`:
```{r}
sp_decorr_preds <- recorrelate_newdata(rf_tpreds, tmoss_test)
```

Finally, we compute our model evaluation metric, root-mean-squared-prediction error (RMSPE) and predictive R-squared (R$^2$) of the test data observations and corresponding predictions:
```{r}
get_rmspe_cor2 <- function(y, preds) {
  errors <- y - preds
  c("rmspe" = sqrt(mean(errors^2)), "cor2" = cor(y, preds)^2)
}
decorr_metrics <- get_rmspe_cor2(moss_test$log_Zn, sp_decorr_preds)
print(decorr_metrics)
```

# Comparing to a Independent (Standard) Random Forest

We can compare the spatial decorrelation transformation to a standard random forest algorithm that assumes independence among observations. First, we must remove the spatial geometry structure from `moss_train` and `moss_test`, as `ranger()` does not accept `sf` objects:
```{r}
moss_train_df <- st_drop_geometry(moss_train)
moss_test_df <- st_drop_geometry(moss_test)
```

Then we may fit the model to training data, predict the test data, and compute the same  RMSPE and R$^2$ metrics:
```{r}
indrfmod <- ranger(form, data = moss_train_df)
indrf_preds <- predict(indrfmod, data = moss_test_df)$predictions
indrf_metrics <- get_rmspe_cor2(moss_test$log_Zn, indrf_preds)
print(indrf_metrics)
```

The spatial decorrelation transformation has a lower RMSPE and higher R$^2$ than the independent random forest, suggesting the spatial decorrelation transformation provides a better fit to the data. This can be visualized is the following observed vs predicted plots of the test data, where the spatial decorrelation predictions tend to more closely adhere to the 1-1 line than the independent random forest predictions:
```{r}
moss_test1 <- moss_test
moss_test1$preds <- sp_decorr_preds
moss_test1$type <- "decorr"
moss_test2 <- moss_test
moss_test2$preds <- indrf_preds
moss_test2$type <- "ind"
moss_test_wide <- rbind(moss_test1, moss_test2)
ggplot(moss_test_wide, aes(x = preds, y = log_Zn, color = type, shape = type)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, color = "orange", linewidth = 2) +
  theme_gray(base_size = 14)
```


# Comparing to a Spatial Residual Random Forest

We can also compare the spatial decorrelation transformation to the spatial residual random forest already available in `spmodel`. The spatial residual random forest fits and independent random forest and the applies a spatial model to the residuals from that independent random forest [@fox2020comparing]. 

```{r}
splmRFmod <- splmRF(form, data = moss_train, spcov_type = "exponential")
splmRF_preds <- predict(splmRFmod, newdata = moss_test)
splmRF_metrics <- get_rmspe_cor2(moss_test$log_Zn, splmRF_preds)
print(splmRF_metrics)
```

The spatial decorrelation transformation outperforms the spatial residual random forest, though the spatial residual random forest still notably outperforms the independent random forest.

# A Grid Search to Estimate the Spatial Decorrelation Transformation Covariance Parameters 

In the prior example, we assumed the true spatial covariance parameters were known. This assumption is unsuitable in practice. One way to estimate the spatial covariance parameters is to use a grid search. In a grid search, several different spatial covariance parameter combinations are evaluated using the spatial decorrelation transformation and the parameter set minimizing some metric (here, test data RMSPE) are the final spatial covariance parameter estimates.
The `decorrelate_grid()` function provides an initial grid of spatial covariance parameters:

```{r}
grid <- decorrelate_grid(
  formula = form,
  data = moss_train,
  spcov_type = "exponential"
)
grid
```

The grid values are chosen to explore different, plausible parts of the parameter space. The magnitudes of the `de` and `ie` variance parameters are based on the sample variance from an independent linear model fit using `lm()`, and the magnitude of the `range` parameters are based on whether or not spatial covariance decays to approximately zero in the spatial domain our outside the spatial domain. 

To estimate which grid values minimize test set RMSPE, we will apply the spatial decorrelation transformation to each row (i.e., parameter set) of the grid and track RMSPE (and R$^2$):

```{r}
y <- moss_test$log_Zn
n_grid <- NROW(grid)
grid$rmspe <- rep(0, n_grid)
grid$cor2 <- rep(0, n_grid)
for (i in seq(1, n_grid)) {
  grid_row <- grid[i, ]
  spcov_params_row <- spcov_params(
    spcov_type = grid_row[["spcov_type"]],
    de = grid_row[["de"]],
    ie = grid_row[["ie"]],
    range = grid_row[["range"]]
  )
  tmoss_train <- decorrelate_data(form, moss_train, spcov_params_row)
  rfmod <- ranger(x = tmoss_train$tX, y = tmoss_train$ty)
  tmoss_test <- decorrelate_newdata(tmoss_train, newdata = moss_test)
  rf_tpreds <- predict(rfmod, data = tmoss_test$tX_newdata)$predictions
  sp_decorr_preds <- recorrelate_newdata(rf_tpreds, tmoss_test)
  errors <- y - sp_decorr_preds
  grid$rmspe[i] <-  sqrt(mean(errors^2))
  grid$cor2[i] <- cor(sp_decorr_preds, y)^2
}
```

The RMSPE and $R^2$ values were added to the grid:
```{r}
grid
```

The estimated spatial covariance parameters are the ones that minimize the test set RMSPE:
```{r}
best_rmspe <- which.min(grid$rmspe)
best_rmspe_row <- grid[best_rmspe, ]
best_rmspe_row
```

We can store these estimated spatial covariance parameters as an `spcov_params()` object to later be used with `decorrelate_data()`, `decorrelate_newdata()`, and `recorrelate_newdata()` to apply the spatial decorrelation transformation:
```{r}
spcov_params_use <- spcov_params(
  spcov_type = best_rmspe_row[["spcov_type"]],
  de = best_rmspe_row[["de"]], 
  ie = best_rmspe_row[["ie"]],
  range = best_rmspe_row[["range"]]
)
spcov_params_use
```

Anisotropy and random effects can also be applied to the grid creation via the `anisotropy` and `random` arguments, whose values can be used later to apply the spatial decorrelation transformation:
```{r}
grid <- decorrelate_grid(
  formula = form,
  data = moss_train,
  spcov_type = "exponential",
  anisotropy = TRUE,
  random = ~ field_rep
)
NROW(grid)
head(grid, n = 5)
```

# Putting it All Together

After finding the spatial covariance parameters that minimize test set RMSPE, we can decorrelate the entire data set and use it make predictions at entirely new locations:
```{r}
tmoss <- decorrelate_data(form, moss, spcov_params = spcov_params_use)
```


# References {.unnumbered}

<div id="refs"></div>
