---
title: "Using emmeans to Estimate Marginal Means of spmodel Objects"
author: "Michael Dumelle, Matt Higham, and Jay M. Ver Hoef"
bibliography: '`r system.file("references.bib", package="spmodel")`'
output:
    html_document:
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
# # jss style
# knitr::opts_chunk$set(prompt=TRUE, echo = TRUE, highlight = FALSE, continue = " + ", comment = "")
# options(replace.assign=TRUE, width=90, prompt="R> ")

# rmd style
knitr::opts_chunk$set(collapse = FALSE, comment = "#>", warning = FALSE, message = FALSE)

# loading
library(ggplot2)
library(spmodel)
library(emmeans)
```

# Introduction {#sec:introduction}

`spmodel` is an \textbf{\textsf{R}} package used to fit, summarize, and
predict for a variety of spatial statistical models. The vignette
provides an estimating marginal means (i.e., least-squares means) of `spmodel` objects using the `emmeans`.
**R** package [@lenth2024emmeans]. Before proceeding, we
load `spmodel` and `emmeans` by running

```{r, eval = FALSE}
library(spmodel)
library(emmeans)
```

If using `spmodel` in a formal publication or report, please cite it.
Citing `spmodel` lets us devote more resources to the package in the
future. We view the `spmodel` citation by running

```{r}
citation(package = "spmodel")
```

# Applying emmeans to spmodel Objects

In this section, we use the point-referenced `lake` data, an `sf` object that contains data on lake conductivty for some southwestern states (Arizona, Colorado, Nevada, Utah) in the United States. We view the first few rows of `lake` by running
```{r}
lake
```

We can learn more about `lake` by running `help("lake", "spmodel")`, and we can visualize the distribution of log conductivity in `lake` by state and year by running
```{r log_cond, fig.cap="Distribution of log conductivity in the lake data.", out.width = "65%", fig.align="center"}
ggplot(lake, aes(color = log_cond)) +
  geom_sf() +
  scale_color_viridis_c() +
  theme_gray(base_size = 14)
```

## A Single-Factor Model

First we explore a single-factor model that characterizes the response variable, log conductivity, by each state (AZ, CO, NV, UT). We fit and summarize this model by running:
```{r}
spmod1 <- splm(
  formula = log_cond ~ state,
  data = lake,
  spcov_type = "exponential"
)
summary(spmod1)
```

The `summary()` output provides mean estimates for each state relative to the difference from a *reference group* (here, AZ). Often, however, the question "What is the mean in each group?" is of interest, and this is not straightforward to obtain from `summary()`. Fortunately, `emmeans` makes this information readily available via the `emmeans` function:
```{r}
em11 <- emmeans(spmod1, ~ state)
```

which, when printed, returns the mean estimates, standard errors, and confidence intervals for each factor level:
```{r}
em11
```

The `emmeans` object `em11` is an `emmGrid` object, but it can easily be converted into a data frame using `data.frame()` or `tibble::as_tibble()`:
```{r}
data.frame(em11)
```

We then visualize the means and confidence intervals:
```{r}
plot(em11)
```

Recall that `summary()` provides mean estimates for each state relative to the difference from a *reference group* (i.e., contrasts with a reference group). Contrasts between mean estimates that are not reference groups, however, is again not straightforward to obtain from `summary()`. Fortunately, `pairs()` provides a simple solution, creating contrasts for comparisons of each factor level to all other factor levels that are easily visualized:
```{r}
pairs(em11)
plot(pairs(em11))
```

By default, the p-values and confidence intervals from the output and plot above are adjusted according to the Tukey method. The model suggests no significant evidence (p-values > 0.1) that the average log conductivity is different among the states. Other p-value adjustment methods can be passed via `adjust`. For example, we can use the Bonferroni method instead of Tukey method 
```{r}
pairs(em11, adjust = "bonferroni")
```

or apply no adjustment method at all:
```{r}
pairs(em11, adjust = "none")
```

## A Multi-Factor Model

Now we explore a model that adds a second factor: `year`, with two levels, 2012 and 2017:
```{r}
spmod2 <- splm(
  formula = log_cond ~ state + year,
  data = lake,
  spcov_type = "exponential"
)
```

We can view the factors separately:
```{r}
em21 <- emmeans(spmod2, ~ state)
em21
em22 <- emmeans(spmod2, ~ year)
em22
```

Or, we can view the factors simultaneously (by providing both variables separated by `+`):
```{r}
em23 <- emmeans(spmod2, ~ state + year)
em23
plot(em23)
```

## A Multi-Factor Model With an Interaction

We can supplement the model with an interaction, which lets the effect of state to vary by year. Recall that shorthand for `state + year + state:year` is `state * year`:
```{r}
spmod3 <- splm(
  formula = log_cond ~ state * year,
  data = lake,
  spcov_type = "exponential"
)
```

Because the effect of state varies by year, single-variable summaries of `emmeans` can be misleading, which `emmeans` warns users about:
```{r, message = TRUE}
emmeans(spmod3, ~ state)
```

Instead, it is helpful to quantify the effect of state *separately for each year*:
```{r}
em31 <- emmeans(spmod2, ~ state, by = "year")
em31
pairs(em31)
plot(em31)
```

And similarly, we can quantify the effect of year *separately for each state*
```{r}
em32 <- emmeans(spmod2, ~ year, by = "state")
pairs(em32)
```

And we can quantify the effect of each combination of `state` and `year`:
```{r}
em33 <- emmeans(spmod2, ~ state + year)
em33
pairs(em33)
```

## A Single-Factor Numeric Model With an Interaction

Suppose it is of interest to supplement the state model (`spmod1`) with a continuous temperature variable:
```{r}
spmod4 <- splm(
  formula = log_cond ~ state * temp,
  data = lake,
  spcov_type = "exponential"
)
```

Because our model has a state-by-year interaction, the effect of temperature varies by state. Supplying the `by` argument lets us quantify the effect of state at the average temperature value:
```{r}
em41 <- emmeans(spmod4, ~ state, by = "temp")
em41
pairs(em41)
```

If we want to quantify the effect of state at specific temperature values, we can supply them via the `at` argument:
```{r}
em42 <- emmeans(spmod4, ~ state, by = "temp", at = list(temp = c(2, 8)))
pairs(em42)
```

We use `emmip()` to visualize the change in the effect of state at varying temperature values:
```{r}
emmip(spmod4, state ~ temp, at = list(temp = c(0:10)), style = "factor")
```

And `emtrends` to quantify the effect of temperature separately for each state:
```{r}
em43 <- emtrends(spmod4, ~ state, var = "temp")
em43
```

## A Multi-Factor Numeric Model With Interactions

We can extend the single-factor numeric model to include multiple factors:
```{r}
spmod5 <- splm(log_cond ~ state * year * temp, data = lake, spcov_type = "exponential")
em51 <- emmeans(spmod5, ~ state, by = c("temp", "year"))
em51
em52 <- emmeans(spmod5, ~ year, by = c("temp", "state"))
em52
em53 <- emmeans(spmod5, ~ state + year, by = "temp")
pairs(em53)
em54 <- emtrends(spmod5, ~ state, by = "year", var = "temp")
em54
em55 <- emtrends(spmod5, ~ year, by = "state", var = "temp")
em55
em56 <- emtrends(spmod5, ~ state + year, var = "temp")
em56
```

# Pairing with `anova()` from `spmodel`

The `anova()` function from `spmodel` is especially helpful to further contextualize `emmeans` output. The `emmeans` functions are very helpful for contrasting factor levels, but `anova()` is built to answer the question "Are any of these factor levels significantly related to the response variable?". Recall `spmod5`, which quantifies the effects of `state`, `year`, and `temp` (and their interactions) on log conductivity:
```{r}
summary(spmod5)
```

We perform an analysis of variance by running:
```{r}
anova(spmod5)
```

The analysis of variance suggests a significant intercept, temperature effect, and state-by-temperature interaction (p-values < 0.01) but no other significant effects (p-values > 0.1).

# Additional Tools

We only showed a small subset of all possible tools that `emmeans` can apply to `spmodel` objects. Additional tools provide support for enhanced visualizations and printing, contrasts, joint hypothesis testing, p-value adjustments, and variable transformations (e.g., when using a `spglm()` or `spgautor()` model object), among many others. To learn more about `emmeans`, visit the website [here](https://rvlenth.github.io/emmeans/).

# References {.unnumbered}

<div id="refs"></div>
