---
title: "Simulating Spatial Data in spmodel"
author: "Michael Dumelle, Matt Higham, and Jay M. Ver Hoef"
bibliography: '`r system.file("references.bib", package="spmodel")`'
output:
  html_document:
editor_options:
  chunk_output_type: console
---

```{r setup, include = FALSE}
# # jss style
# knitr::opts_chunk$set(prompt=TRUE, echo = TRUE, highlight = FALSE, continue = " + ", comment = "")
# options(replace.assign=TRUE, width=90, prompt="R> ")

# rmd style
knitr::opts_chunk$set(collapse = FALSE, comment = "#>", warning = FALSE, message = FALSE)

# loading
library(spmodel)
library(ggplot2)
```

# Introduction {#sec:introduction}

`spmodel` is an **R** package used to fit, summarize, and
predict for a variety of spatial statistical models. This vignette explores `sprnorm()`, which simulates spatially correlated normal (Gaussian) random variables, along with its companions for simulating non-Gaussian spatial data. Before proceeding, we
load `spmodel` and `ggplot2` by running
```{r, eval = FALSE}
library(spmodel)
library(ggplot2)
```

If using `spmodel` in a formal publication or report, please cite it. Citing `spmodel` lets us devote more resources to it in the future. We view the `spmodel` citation by running

```{r}
citation(package = "spmodel")
```

Before proceeding, we set a reproducible seed:
```{r}
set.seed(0)
```


# Simulating Spatially Correlated Data

`sprnorm()` is `spmodel`'s spatial analogue to base-**R**'s `rnorm()` for simulating spatially correlated normal (i.e., Gaussian) data. `sprnorm()` requires an [spcov_params()] object that specifies the (known) spatial covariance parameters and a `data` argument with the locations to simulate at. We define an exponential spatial covariance with dependent variance (`de`) of 2, independent variance (`ie`) of 0.1, and range of 2, and simulate 3,000 locations completely at random on the unit square:
```{r}
n <- 3000
sim_dat <- data.frame(xc = runif(n), yc = runif(n))

params <- spcov_params(
  spcov_type = "exponential",
  de = 2,
  ie = 0.1,
  range = 2
)
```

We simulate a single spatially correlated realization by running
```{r}
sim_dat$val <- sprnorm(params, data = sim_dat, xcoord = xc, ycoord = yc)
ggplot(sim_dat, aes(x = xc, y = yc, color = val)) +
  geom_point(size = 1) +
  scale_color_viridis_c() +
  theme_bw(base_size = 14)
```

`sprnorm()` simulates data with mean zero by default; a non-constant mean can be supplied via the `mean` argument. Multiple independent realizations can be simulated at once via the `samples` argument, which returns a matrix (one column per sample) instead of a vector. Generating many samples via `samples` this way costs nearly the same (computationally) as generating a single sample, as the primary cost of factoring an $n \times n$ covariance matrix occurs only once and is reused for each sample. For technical details regarding simulating data in `spmodel`, see the Technical Details vignette.

# Simulating Non-Gaussian Data

`spmodel` also provides `sprpois()`, `sprnbinom()`, `sprbinom()`, `sprbeta()`, `sprgamma()`, and `sprinvgauss()` for simulating other spatially correlated random variables (Poisson, negative binomial, binomial, beta, gamma, and inverse Gaussian, respectively). Each works by first simulating a latent process with `sprnorm()` and then simulating the requested distribution conditional on that latent process.

We simulate a single spatially correlated realization of binomial data by running
```{r}
n <- 3000
sim_dat <- data.frame(xc = runif(n), yc = runif(n))
sim_dat$val <- as.factor(sprbinom(params, data = sim_dat, xcoord = xc, ycoord = yc))
ggplot(sim_dat, aes(x = xc, y = yc, color = val)) +
  geom_point(size = 1) +
  scale_color_manual(values = c("blue", "orange")) +
  theme_bw(base_size = 14)
```

# References {.unnumbered}

<div id="refs"></div>
