---
title: "Spatial Generalized Linear Models in spmodel"
author: "Michael Dumelle, Matt Higham, and Jay M. Ver Hoef"
header-includes:
   - \usepackage{amsmath,amsfonts,amssymb}
   - \usepackage{bm, bbm}
   - \usepackage{mathtools}
bibliography: '`r system.file("references.bib", package="spmodel")`'
output:
    pdf_document:
      number_sections: true
      toc: false
pkgdown:
  as_is: true
---

```{r setup, include = FALSE}
# # jss style
# knitr::opts_chunk$set(prompt=TRUE, echo = TRUE, highlight = FALSE, continue = " + ", comment = "")
# options(replace.assign=TRUE, width=90, prompt="R> ")

# rmd style
knitr::opts_chunk$set(collapse = FALSE, comment = "#>", warning = FALSE, message = FALSE)

# loading
library(ggplot2)
library(spmodel)
```

# Introduction {#sec:introduction}


`spmodel` is an \textbf{\textsf{R}} package used to fit, summarize, and predict for a variety of spatial statistical models. The vignette provides an introduction spatial generalized linear models for non-Gaussian response distributions in `spmodel`. Before proceeding, we load `spmodel` by running
```{r, eval = FALSE}
library(spmodel)
```

If using `spmodel` in a formal publication or report, please cite it. Citing `spmodel` lets us devote more resources to the package in the future. We view the `spmodel` citation by running
```{r}
citation(package = "spmodel")
```

We will create visualizations using `ggplot2` [@wickham2016ggplot2], which we load by running
```{r, eval = FALSE}
library(ggplot2)
```

`ggplot2` is only installed alongside `spmodel` when `dependencies = TRUE` in `install.packages()`, so check that the package is installed and loaded before reproducing any of these vignette's visualizations. We will also show code that can be used to create interactive visualizations of spatial data with `mapview` [@appelhans2022mapview]. `mapview` also has many backgrounds available that contextualize spatial data with topographical information. Before running the `mapview` code interactively, make sure `mapview` is installed and loaded.

`spmodel` contains various methods for generic functions defined outside of `spmodel`. To find relevant documentation for these methods, run `help("generic.spmodel", "spmodel")` (e.g., `help("summary.spmodel", "spmodel")`, `help("predict.spmodel", "spmodel")`, etc.).
Note that `?generic.spmodel` is shorthand for `help("generic.spmodel", "spmodel")`. We provide more details and examples regarding these methods and generics throughout this vignette. For a full list of `spmodel` functions available, see `spmodel`'s documentation manual.

# A Review of the Spatial Linear Model

We first focus on spatial linear models for a Gaussian response vector. The spatial linear model is defined as
\begin{equation}\label{eq:splm}
 \mathbf{y} = \mathbf{X} \boldsymbol{\beta} + \boldsymbol{\tau} + \boldsymbol{\epsilon},
\end{equation}
where for a sample size $n$, $\mathbf{y}$ is an $n \times 1$ column vector of response variables, $\mathbf{X}$ is an $n \times p$ design (model) matrix of explanatory variables, $\boldsymbol{\beta}$ is a $p \times 1$ column vector of fixed effects controlling the impact of $\mathbf{X}$ on $\mathbf{y}$, $\boldsymbol{\tau}$ is an $n \times 1$ column vector of spatially dependent random errors, and $\boldsymbol{\epsilon}$ is an $n \times 1$ column vector of spatially independent random errors. We make a few assumptions about $\boldsymbol{\tau}$ and $\boldsymbol{\epsilon}$ in Equation$~$\ref{eq:splm}: first, that $\text{E}(\boldsymbol{\tau}) = \text{E}(\boldsymbol{\epsilon}) = \boldsymbol{0}$, where $\text{E}(\cdot)$ denotes expectation; second, that $\text{Cov}(\boldsymbol{\tau}) = \sigma^2_\tau \mathbf{R}$, where $\mathbf{R}$ is an $n \times n$ matrix that determines the spatial dependence structure in $\mathbf{y}$ and depends on a range parameter, $\phi$; third, that $\text{Cov}(\boldsymbol{\epsilon}) = \sigma^2_\epsilon \mathbf{I}$, where $\mathbf{I}$ is an $n \times n$ identity matrix; and fourth, that $\boldsymbol{\tau}$ and $\boldsymbol{\epsilon}$ are independent of one another. The parameter $\sigma^2_{\tau}$ is called the spatially dependent random error variance or partial sill. The parameter $\sigma^2_\epsilon$ is called the spatially independent random error variance or nugget. These two variance parameters are henceforth more intuitively written as $\sigma^2_{de}$ and $\sigma^2_{ie}$, respectively.  The covariance of $\mathbf{y}$ is denoted $\boldsymbol{\Sigma}$ and given by
\begin{equation}\label{eq:spcov}
 \boldsymbol{\Sigma} = \sigma^2_{de}\mathbf{R} + \sigma^2_{ie} \mathbf{I}.
\end{equation}
The parameters $\sigma^2_{de}$, $\phi$, and $\sigma^2_{ie}$ are elements of $\boldsymbol{\theta}$, the covariance parameter vector.

The spatial linear model applies to both point-referenced and areal (i.e., lattice; polygon) data. Spatial data are point-referenced when the elements in $\mathbf{y}$ are observed at point-locations indexed by x-coordinates and y-coordinates on a spatially continuous surface with an infinite number of locations. For example, consider sampling soil at any point-location in a field.  Spatial linear models for point-referenced data are sometimes called geostatistical models. An example of an $\mathbf{R}$ matrix for point-referenced data is the spherical correlation matrix given by
\begin{equation}\label{eq:spherical}
\mathbf{R} = \left(1 - 1.5\frac{\mathbf{H}}{\phi} + 0.5\frac{\mathbf{H}^3}{\phi^3}\right) \odot \mathcal{I}(\mathbf{H} \leq \phi),
\end{equation}  
where $\mathbf{H}$ is a matrix of Euclidean distances among observations, $\mathbf{H}^3$ is a matrix of cubed Euclidean distances,  $\odot$ is the Hadmard (element-wise; direct) product, and $\mathcal{I}(\mathbf{H} \leq \phi)$ is an indicator function equal to one when the $ij$th element of $\mathbf{H}$ is less than $\phi$ (the range) and zero otherwise.  A second is example is the Matérn correlation matrix given by
\begin{equation}\label{eq:matern}
\mathbf{R} = \frac{2^{1 - \nu}}{\Gamma(\nu)} \left(2\nu \mathbf{H} / \phi \right)^{\nu / 2} \odot B_k(2\nu \mathbf{H}, \nu),
\end{equation}
where $\nu$ is a smoothness parameter, $\odot$ is the Hadmart (element-wise; direct) product, and $B_k(\cdot, \nu)$ is a Bessel-K function with order $\nu$. @chiles2012geostatistics provide examples of several spatial dependence structures.

Spatial data are areal when they are observed as part of a finite network of polygons whose connections are indexed by a neighborhood structure. For example, the polygons may represent counties in a state who are neighbors if they share at least one boundary. An example of an $\mathbf{R}$ matrix for areal data is the simultaneous-autoregressive spatial correlation matrix given by
\begin{equation*}
  \mathbf{R} = [(\mathbf{I} - \phi \mathbf{W})(\mathbf{I} - \phi \mathbf{W})^\top]^{-1},
\end{equation*}
where $\mathbf{W}$ is a weight matrix that describes the neighborhood structure among observations. Spatial linear models for areal data are sometimes called spatial autoregressive models. For thorough reviews of the spatial linear model, see @cressie1993statistics, @banerjee2014hierarchical, and @schabenberger2017statistical.

# The Spatial Generalized Linear Model

It is common to implicitly assume $\mathbf{y}$ in the spatial linear model is unconstrained and Gaussian. This is a restrictive assumption and often inappropriate for binary data, proportion data, count data, and skewed data, which are common in many practical applications. These types of data are most naturally described using generalized linear models [@nelder1972generalized; @mccullagh1989generalized]. Generalized linear models are an extension of linear models that "link" a function of a distribution's mean to the linear function $\mathbf{X} \boldsymbol{\beta}$  (the explanatory variables and fixed effects). We can extend generalized linear models by adding linear, latent (i.e., unobserved) random effects to $\mathbf{X} \boldsymbol{\beta}$. These types of models are called generalized linear mixed models [@breslow1993approximate] and are often formulated using a hierarchical construction [@lee1996hierarchical]. For thorough reviews of generalized linear models and generalized linear mixed models, see @bolker2009generalized, @myers2012generalized, @faraway2016extending, and @wood2017generalized.

A recent `spmodel` release (version 0.4.0) provided support for spatial generalized linear models (i.e., a generalized linear mixed model with spatial random effects). We define the spatial generalized linear model as
\begin{equation}\label{eq:spglm}
f(\boldsymbol{\mu}) \equiv \mathbf{w} = \mathbf{X} \boldsymbol{\beta} + \boldsymbol{\tau} + \boldsymbol{\epsilon},
\end{equation}
where $f(\cdot)$ is the link function, $\boldsymbol{\mu}$ is the mean of $\mathbf{y}$, and the remaining terms $\mathbf{X}$, $\boldsymbol{\beta}$, $\boldsymbol{\tau}$, $\boldsymbol{\epsilon}$ represent the same quantities as for spatial linear models in Equation$~$\ref{eq:splm}. That is, $\text{E}(\boldsymbol{\tau}) = \text{E}(\boldsymbol{\epsilon}) = \boldsymbol{0}$, $\text{Cov}(\boldsymbol{\tau}) = \sigma^2_{de}\mathbf{R}$, $\text{Cov}(\boldsymbol{\epsilon}) = \sigma^2_{ie} \mathbf{I}$, $\boldsymbol{\tau}$ is independent of $\boldsymbol{\epsilon}$, and $\boldsymbol{\Sigma} = \sigma^2_{de} \mathbf{R} + \sigma^2_{ie} \mathbf{I}$. The link function, $f(\cdot)$, "links" a function of $\boldsymbol{\mu}$ to the linear term $\mathbf{w}$. Note that the linking of $\boldsymbol{\mu}$ to $\mathbf{w}$ applies element-wise to each vector. Each link function $f(\cdot)$ has a corresponding inverse link function, $f^{-1}(\cdot)$, which "links" a function of $\mathbf{w}$ to $\boldsymbol{\mu}$.  Notice that $\mathbf{w}$ is unconstrained but $\boldsymbol{\mu}$ is usually constrained in some way (e.g., positive) that depends on the distribution assumed for $\mathbf{y}$. 

Spatial linear and generalized linear models rely on the same components: explanatory variables, fixed effects, spatially dependent random errors, and spatial independent random errors. These components are related to the response linearly (spatial linear models) or non-linearly (spatial generalized linear models). The response itself is generally assumed to follow a Gaussian distribution (spatial linear models) or one of many non-Gaussian distributions (spatial generalized linear models). Some non-Gaussian distributions include the binomial distribution for binary responses, the beta distribution for proportion responses, the Poisson or negative binomial distributions for count responses, and the gamma or inverse Gaussian distributions for skewed, positive continuous responses.

`spmodel` implements a marginal approach to inference that uses a novel application of the Laplace approximation. This marginal approach is powerful and flexible, accommodating a wide range of covariance structures. It is a frequentist approach, which means that it does not rely on any Bayesian sampling routines, as is common when handling complex hierarchical structures. It also formally maximizes a likelihood, which is advantageous because it means that likelihood-based statistics such as AIC [@akaike1974new; @hoeting2006model] are defined, in contrast to approaches that only specify the first two moments of a distribution (e.g., quasi-likelihood fit via iteratively re-weighted least squares). @ver2023marginal provide the methodology's full details. Table$~$\ref{table:links}.

`spmodel` accommodates the binomial, beta, Poisson, negative binomail, gamma, and inverse Gaussian distributions. The binomial and beta response distributions (specified by \texttt{family}) use the logit link, while the Poisson, negative binomial, gamma, and inverse Gaussian distributions use the log link (Table$~$\ref{table:links}).

\begin{table}
\centering
\begin{tabular}{||c|ccc||} 
 \hline
 Family & Link Function & Link Name & Data Type  \\ [0.5ex] 
 \hline \hline
 Binomial & $f(\mathbf{\mu}) = \log(\mathbf{\mu} / (1 - \mathbf{\mu}))$ & Logit & Binary; Binary Count \\
 Poisson & $f(\mathbf{\mu}) = \log(\mathbf{\mu})$ & Log & Count \\
 Negative Binomial & $f(\mathbf{\mu}) = \log(\mathbf{\mu})$ & Log & Count \\
 Beta & $f(\mathbf{\mu}) = \log(\mathbf{\mu} / (1 - \mathbf{\mu}))$ & Logit & Proportion \\
 Gamma & $f(\mathbf{\mu}) = \log(\mathbf{\mu})$ & Log & Skewed \\
 Inverse Gaussian & $f(\mathbf{\mu}) = \log(\mathbf{\mu})$ & Log & Skewed \\ 
 \hline\hline
\end{tabular}
\caption{Spatial generalized linear model response distributions available in `spmodel` alongside their corresponding link functions and data types.}
\label{table:links}
\end{table}

## Model fitting {#sec:model-fit}

Spatial generalized linear models in `spmodel` are fit using \texttt{spglm()} (for point-referenced data) or \texttt{spgautor()} (for areal data). These functions are similar in structure both to one another and to the \texttt{glm()} function in base **R**. They generally require the following four arguments: \texttt{formula}, a formula that describes the relationship between the response and explanatory variables; \texttt{family}, the response distribution, which can be \texttt{binomial}, \texttt{beta}, \texttt{poisson}, \texttt{nbinomial}, \texttt{Gamma}, or \texttt{inverse.gaussian}; \texttt{data}, a \texttt{data.frame} or \texttt{sf} object [@pebesma2018sf] whose rows index observations and whose columns contain the response variable, explanatory variables, and spatial information; and \texttt{spcov\_type}, one of 19 spatial covariance types (e.g., spherical spatial covariance, Matérn spatial covariance, simultaneous-autoregressive spatial covariance etc.). The binomial and beta response distributions (specified by \texttt{family}) use the logit link, while the Poisson, negative binomial, gamma, and inverse Gaussian distributions use the log link (Table$~$\ref{table:links}). Other arguments to \texttt{spglm()} and \texttt{spgautor()} are either required only in specific situations or have default values specified. All arguments to the \texttt{spglm()} and \texttt{spgautor()} functions are summarized in Table$~$\ref{table:model-fit}. Models for several hundred observations can take anywhere from a few seconds to a few minutes to fit, depending on the shape of the likelihood surface and how quickly the covariance parameters converge. Models for several thousand observations can utilize `spmodel`'s large data set options (Table$~$\ref{table:model-fit}).
\begin{table}
\centering
\begin{tabular}{||c|c|c||} 
 \hline
 Argument & Purpose & Applies To \\ [0.5ex] 
 \hline\hline
 \texttt{formula} & Fixed effects formula & Both  \\ 
 \texttt{family} & Response distribution & Both  \\
 \texttt{data} & Data Frame or \texttt{sf} object & Both  \\
 \texttt{spcov\_type} & Spatial covariance type & Both  \\
 \texttt{spcov\_initial} & Initial or known spatial covariance parameters & Both  \\ 
 \texttt{dispersion\_initial} & Initial or known dispersion parameter & Both  \\
 \texttt{estmethod} & Estimation method & Both  \\
 \texttt{random} & Random effects formula & Both  \\
 \texttt{randcov\_initial} & Initial or known random effect variances & Both  \\ 
 \texttt{partition\_factor} & A partition factor formula & Both  \\
 \texttt{...} & Additional arguments & Both \\
 \hline \hline
 \texttt{anisotropy} & Whether to model anisotropy & \texttt{spglm()} \\
 \texttt{xcoord} & An x-coordinate name & \texttt{spglm()}  \\
 \texttt{ycoord} & A y-coordinate name & \texttt{spglm()}  \\
 \texttt{local} & Options for large data sets & \texttt{spglm()}  \\
 \hline \hline
 \texttt{W} & A neighborhood weight matrix & \texttt{spgautor()}  \\
 \texttt{row\_st} & Whether to row-standardize \texttt{W} & \texttt{spgautor()}  \\
 \texttt{M} & A symmetry matrix & \texttt{spgautor()}  \\
 \hline
\end{tabular}
\caption{Arguments to \texttt{spglm()} and \texttt{spgautor()}, their purpose, and whether the argument applies to both \texttt{spglm()} and \texttt{spgautor()} or just one of them.}
\label{table:model-fit}
\end{table}

## Model evaluation and diagnostics {#sec:diagnostics}

After fitting a model, it is usually beneficial to evaluate certain aspects of model fit or inspect and visualize model diagnostics like residuals. There are many functions in `spmodel` that are used for these purposes, and their main argument is the fitted model object returned by \texttt{spglm()} or \texttt{spgautor()}. The names of these functions are meant to be illustrative, describing the purpose of the function itself. The functions for evaluating model fit (Table$~$\ref{table:model-evaluation}) return likelihood-based statistics like AIC, AICc, and deviance [@mccullagh1989generalized], return coefficient estimates and confidence intervals, perform cross validation, and more. The functions for inspecting model diagnostics (Table$~$\ref{table:model-diagnostics}) return Cook's distances [@cook1979influential], fitted values, leverage (hat) values, residuals (of several types), and visualizations.

\begin{table}
\centering
\begin{tabular}{||c|c||} 
 \hline
 Function & Purpose \\ [0.5ex] 
 \hline\hline
 \texttt{AIC()}; \texttt{AICc()} & Return AIC and AICc   \\ 
 \texttt{anova()} & Perform an analysis of variance \\
 \texttt{coef()}; \texttt{coefficients()} & Return parameter estimates   \\
 \texttt{confint()} & Return confidence intervals  \\ 
 \texttt{covmatrix()} & Return the fitted covariance matrix   \\
 \texttt{deviance()} & Return the deviance  \\
 \texttt{glance()}; \texttt{glances()} & Glance at model fits \\ 
 \texttt{logLik()} & Return the log-likelihood  \\
 \texttt{loocv()} & Perform leave-one-out cross validation \\
 \texttt{pseudoR2()} & Return the pseudo R-squared   \\
 \texttt{summary()} & Summarize the fitted model   \\
 \texttt{tidy()} & Tidy the fitted model   \\
 \texttt{varcomp()} & Compare sources of variability \\
 \texttt{vcov()} & Return variance-covariance matrices  \\
 \hline
\end{tabular}
\caption{Model evaluation functions available for models fit using \texttt{spglm()} or \texttt{spgautor()} and their purpose.}
\label{table:model-evaluation}
\end{table}

\begin{table}
\centering
\begin{tabular}{||c|c||} 
 \hline
 Function & Purpose \\ [0.5ex] 
 \hline\hline
 \texttt{augment()} & Augment model data with diagnostics \\
 \texttt{cooks.distance()} & Return Cook's distances   \\ 
 \texttt{fitted()}; \texttt{fitted.values()} & Return fitted values   \\
 \texttt{hatvalues()}; & Return hat (leverage) values \\
 \texttt{plot()} & Plot model diagnostics   \\
 \texttt{residuals()}; \texttt{resid()}; \texttt{rstandard()} & Return residuals  \\ 
 \hline
\end{tabular}
\caption{Model diagnostic functions available for models fit using \texttt{spglm()} or \texttt{spgautor()} and their purpose.}
\label{table:model-diagnostics}
\end{table}

The functions \texttt{tidy()}, \texttt{glance()}, and \texttt{augment()}, popularized by the `broom` **R** package [@robinson2021broom], are particularly useful. The \texttt{tidy()} function tidies parameter estimate output, returning a special \texttt{data.frame} called a \texttt{tibble} [@muller2021tibble] that is much easier to manipulate and work with than the parameter estimate output printed by \texttt{summary()} to the **R** console. The \texttt{glance()} function glances at the model fit, returning a \texttt{tibble} with the sample size, number of estimated parameters, and several model-fit statistics. \texttt{glances()} is an extension of \texttt{glance()} that glances at multiple models simultaneously. The \texttt{augment()} function augments the model fit, returning a \texttt{tibble} with the data used to fit the model as well as model diagnostics. The \texttt{augment()} function can also be used to augment prediction data, as we discuss next.

## Prediction {#sec:prediction}

Often a priority of a spatial data analysis is using the fitted model to make predictions at unobserved locations and quantify uncertainties in those predictions. The \texttt{predict()} function in `spmodel` is used to predict the mean process at unobserved locations. It is similar in structure to \texttt{predict()} for objects fit using \texttt{glm()} in base **R** and takes two main arguments: \texttt{object}, the fitted model object returned by \texttt{spglm()} or \texttt{spgautor()}; and \texttt{newdata}, a \texttt{data.frame} or \texttt{sf} object whose rows index locations requiring prediction and whose columns contain the explanatory variables and spatial information at these locations. Other arguments to \texttt{predict()} are either required only in specific situations or have default values specified. All arguments to the \texttt{predict()} function are summarized in Table$~$\ref{table:prediction}. The \texttt{augment()} function can also be used to augment \texttt{newdata} with predictions, taking the same arguments as \texttt{predict()}.

\begin{table}
\centering
\begin{tabular}{||c|c||} 
 \hline
 Argument & Purpose \\ [0.5ex] 
 \hline\hline
 \texttt{object} & Fitted model object \\
 \texttt{newdata} & New data frame or \texttt{sf} object   \\
 \texttt{type} & Whether predictions are on the link or response scale \\
 \texttt{se.fit} & Whether standard errors should be returned   \\
 \texttt{interval} & Interval type (prediction or confidence) \\
 \texttt{level} & Prediction or confidence level   \\
 \texttt{local} & Options for large data sets  \\ 
 \texttt{var\_correct} & Whether prediction variances should be adjusted \\
 \texttt{...} & Additional arguments \\
 \hline
\end{tabular}
\caption{Arguments to \texttt{predict()} and their purpose.}
\label{table:prediction}
\end{table}

## Simulation {#sec:simulation}

`spmodel` has functions (Table$~$\ref{table:simulation}) that are used to simulate spatial data from Gaussian, binomial, beta, Poisson, negative binomial, gamma, and inverse Gaussian response distributions. These simulation functions are similar in structure to commonly used base **R** functions for simulating data. They are also similar in name, simply adding an \texttt{sp} prefix to the conventional name in base **R**. For example, \texttt{rpois()} is used to simulate Poisson data in base **R**, while \texttt{sprpois()} is used to simulate spatial Poisson data in `spmodel`.

\begin{table}
\centering
\begin{tabular}{||c|c||} 
 \hline
 Function & Purpose \\ [0.5ex] 
 \hline\hline
 \texttt{sprnorm()} & Simulate spatial Gaussian data \\
 \texttt{sprbinom()} & Simulate spatial binomial data \\
 \texttt{sprbeta()} & Simulate spatial beta data \\
 \texttt{sprpois()} & Simulate spatial Poisson data \\
 \texttt{sprnbinom()} & Simulate spatial negative binomial data \\
 \texttt{sprgamma()} & Simulate spatial gamma data \\
 \texttt{sprinvgauss()} & Simulate spatial inverse Gaussian data \\
 \hline
\end{tabular}
\caption{Functions to simulate spatial data and their purpose.}
\label{table:simulation}
\end{table}

# An application to binary (presence/abscence) data {#sec:application-binary}

In the next two sections we to show how to use `spmodel` to analyze data collected on moose in the Togiak region of Alaska, USA. The \texttt{moose} data in `spmodel` contain information on moose counts from a survey of sites in the Togiak region. The data are stored as an `sf` object, a special \texttt{data.frame} built to store spatial data. \texttt{moose} has several variables: \texttt{strat}; a stratification variable with two levels (\texttt{L} for low and \texttt{M} for medium) based on surrounding landscape metrics; \texttt{elev}, the site elevation; \texttt{presence}, whether or not at least one moose was observed at the site; \texttt{count}, the number of moose observed at a the site; and \texttt{geometry}, the spatial coordinates in an Alaska Albers projection (EPSG: 3338). We load `spmodel` and the \texttt{moose} data by running
```{r}
library("spmodel")
data("moose")
```

We are interested in quantifying the effects of \texttt{elev} and \texttt{strat} on moose \texttt{presence} and begin with some visualizations. First, we visualize the presence for observed sites:

```{r}
ggplot(moose, aes(color = presence)) + 
  geom_sf(size = 2) + 
  scale_color_viridis_d(option = "H") +
  theme_gray(base_size = 14)
```

There is spatial patterning in the distribution of moose presence, as moose tend to be present in the eastern and southwestern parts of the domain. This spatial pattern seems to have a directional orientation, seemingly strongest in the northwest-to-southeast direction.

Next, we visualize the elevation for each strata:
```{r}
ggplot(moose, aes(color = elev)) +
  geom_sf(size = 1) + 
  facet_grid(~ strat) +
  scale_color_viridis_c()  +
  scale_y_continuous(breaks = seq(62.6, 63.6, length.out = 5)) +
  scale_x_continuous(breaks = seq(-148.5, -146, length.out = 3)) +
  theme_gray(base_size = 12)
```

There are no obvious spatial patterns in the assignment of sites to strata, while elevation is highest on the boundaries of the domain and lowest in the middle of the domain. There is spatial patterning in the distribution of moose presence, as moose tend to be present in the eastern and southwestern parts of the domain. This spatial pattern seems to have a directional orientation, seemingly strongest in the northwest-to-southeast direction.

To quantify the effects of \texttt{elev}, \texttt{strat}, and spatial dependence on moose presence, we fit three binomial (i.e., ) logistic regression models using \texttt{spglm()}:
```{r}
# model one
bin_mod <- spglm(
  formula = presence ~ elev + strat + elev:strat,
  family = "binomial",
  data = moose,
  spcov_type = "none"
)
# model two
bin_spmod <- spglm(
  formula = presence ~ elev + strat + elev:strat,
  family = "binomial",
  data = moose,
  spcov_type = "spherical"
)
# model three
bin_spmod_anis <- spglm(
  formula = presence ~ elev + strat + elev:strat,
  family = "binomial",
  data = moose,
  spcov_type = "spherical",
  anisotropy = TRUE
)
```

All three models have the same fixed effect structure, using elevation, strata, and their interaction to explain moose presence. The three models vary in their covariance structure. The first model, \texttt{bin\_mod}, has no spatial covariance ($\sigma^2_{de} = 0$ in Equation$~$\ref{eq:spcov}). The second model, \texttt{bin\_spmod} has a spherical spatial covariance (Equation$~$\ref{eq:spherical}). The third model, \texttt{bin\_spmod\_anis}, has a spherical spatial covariance that incorporates geometric anisotropy. Geometric anisotropy allows the spatial covariance to vary with direction by evaluating Equation$~$\ref{eq:spherical} with $\mathbf{H}^*$ (instead of $\mathbf{H}$), a matrix of distances between transformed coordinates that are rotated and scaled appropriately [@schabenberger2017statistical]. We evaluate the three models using AIC by running
```{r}
AIC(bin_mod, bin_spmod, bin_spmod_anis)
```
The spatial models (\texttt{bin\_spmod}, \texttt{bin\_spmod\_anis}) have a much lower AIC than the non-spatial model (\texttt{bin\_mod}), which suggests that the models benefit from incorporating spatial dependence. \texttt{bin\_spmod\_anis} has a lower AIC than \texttt{bin\_spmod}, which suggests that the model benefits from incorporating directionality in the spatial dependence. Next we inspect \texttt{bin\_spmod\_anis} and later use it to predict moose presence probability at unobserved sites.

We summarize \texttt{bin\_spmod\_anis} using \texttt{summary()}:
```{r}
summary(bin_spmod_anis)
```
The \texttt{summary()} output from an \texttt{spglm()} model is very similar to the \texttt{summary()} output from a base-**R** \texttt{glm()} model, returning the original function call, deviance residuals, a fixed effects coefficients table, a pseudo R-squared (which quantifies the amount of variability explained by the fixed effects), and covariance parameter coefficients. The fixed effects coefficients table provides some evidence that elevation and strata are associated with with moose presence and that the effect of elevation varies across the two strata (all $p~$values $<$ 0.05). The \texttt{plot()} function returns graphics of several model diagnostics. Running
```{r, eval = FALSE}
plot(bin_spmod_anis, which = c(7, 8))
```

```{r, fig.show='hold', out.width="50%"}
plot(bin_spmod_anis, which = 7)
plot(bin_spmod_anis, which = 8)
```

shows that the spatial dependence is evident and appears strongest in the northwest-to-southeast direction. These findings are supported by the clear spatial patterns in the data.

Recall that the \texttt{tidy()}, \texttt{glance()}, and \texttt{augment()} functions are particularly useful tools for manipulating and understanding fitted model objects. The \texttt{tidy()} function tidies model output, returning a \texttt{tibble} of parameter estimates (and confidence intervals):
```{r}
tidy(bin_spmod_anis, conf.int = TRUE)
```
By default, the fixed effects estimates are returned, but covariance parameter estimates are returned via the \texttt{effects} argument:
```{r}
tidy(bin_spmod_anis, effects = "spcov")
```
The \texttt{is\_known} column indicates whether covariance parameters were assumed known, which is possible to specify using the \texttt{spcov\_initial} argument to \texttt{spglm()}. Here, all parameters were assumed unknown and then estimated.

The \texttt{glance()} function glances at model fit, returning a \texttt{tibble} of model fit statistics:
```{r}
glance(bin_spmod_anis)
```
\texttt{glances()} can be used to glance at multiple models simultaneously and sorts models by ascending AICc:
```{r}
glances(bin_mod, bin_spmod, bin_spmod_anis)
```

The \texttt{augment()} function augments the data with model diagnostics:
```{r}
aug_mod <- augment(bin_spmod_anis)
aug_mod
```
where \texttt{.fitted} are fitted values (the estimated $f^{-1}(\mathbf{w})$, or moose presence probability), \texttt{.resid} are response residuals, \texttt{.hat} values indicate leverage, \texttt{.cooksd} values indicate Cook's distance (i.e., influence), and \texttt{.std.resid} are standardized residuals.

The \texttt{moose\_preds} data contains survey sites that were not sampled. We can use \texttt{bin\_spmod\_anis} to make predictions of the underlying probabilities of moose presence at these sites using \texttt{predict()} or \texttt{augment()}. \texttt{predict()} and \texttt{augment()} return the same predictions but \texttt{augment()} augments the prediction data with the predictions:
```{r}
data("moose_preds")

spmod_anis_preds <- predict(
  bin_spmod_anis,
  newdata = moose_preds,
  type = "response",
  interval = "prediction"
)
head(spmod_anis_preds)

aug_pred <- augment(
  bin_spmod_anis,
  newdata = moose_preds,
  type = "response", 
  interval = "prediction"
)
aug_pred
```

Estimated moose presence probability for the \texttt{moose} data (obtained via \texttt{fitted()}) and predicted moose presence probability for the \texttt{moose\_preds} data (obtained via \texttt{predict()} or \texttt{augment()}) are overlain below and share similar patterns:

```{r}
aug_mod$type <- "mod"
aug_pred$type <- "pred"
keep_cols <- c(".fitted", "type")
aug_combined <- rbind(aug_mod[, keep_cols], aug_pred[, keep_cols])
ggplot(aug_combined, aes(color = .fitted, shape = type)) +
  geom_sf(size = 2) +
  scale_color_viridis_c(option = "H")  +
  theme_gray(base_size = 14)
```

# An application to count data {#sec:application-count}

The \texttt{moose} data also contain the number of moose observed at each site:

```{r}
ggplot(moose, aes(color = count)) +
  geom_sf(size = 2) +
  scale_color_viridis_c(option = "H")  +
  theme_gray(base_size = 14)
```


The moose counts are similarly distributed as moose presence, highest in the eastern and southwestern parts of the domain. We compare two count models, Poisson and negative binomial. The Poisson model assumes the underlying process generating the counts has the same mean and variance while the negative binomial model allows for overdispersion (i.e., the variance is greater than the mean). We are interested in quantifying the effects of \texttt{elev} and \texttt{strat} on moose `count`, albeit using a slightly different approach than we did for moose `presence`. Here, we will model elevation as a fixed effect and allow this effect to change between strata, but we will model strata as a random effect to highlight additional flexibility of the `spmodel` package. We fit relevant Poisson and negative binomial models with a Matérn spatial covariance (Equation$~$\ref{eq:matern}) using \texttt{spglm()}:

```{r}
count_mod_pois <- spglm(
  count ~ elev + elev:strat,
  family = "poisson",
  data = moose,
  spcov_type = "matern",
  random = ~ (1 | strat)
)

count_mod_nb <- spglm(
  count ~ elev + elev:strat,
  family = "nbinomial",
  data = moose,
  spcov_type = "matern",
  random = ~ (1 | strat)
)
```

Random effects are specified in `spmodel` via the \texttt{random} argument using similar syntax as the commonly used `lme4` [@bates2015lme4] and `nlme` [@pinheiro2006mixed] **R** packages for non-spatial mixed models. In `spmodel`, \texttt{~ strat} is short-hand for \texttt{~ (1 | strat)}. Random effects alter the covariance structure of the model, building additional correlation into the model for sites sharing a level of the random effect (here, sites within the same strata). More formally, when incorporating a random effect, Equation$~$\ref{eq:spglm} becomes 
\begin{equation}\label{eq:spglm-rand}
f(\boldsymbol{\mu}) \equiv \mathbf{w} = \mathbf{X} \boldsymbol{\beta} + \mathbf{Z}\mathbf{v} + \boldsymbol{\tau} + \boldsymbol{\epsilon},
\end{equation}
where $\mathbf{Z}$ is a design matrix that indexes the random effects, $\mathbf{v}$, and $\text{Cov}(\mathbf{v}) = \sigma^2_{v}\mathbf{I}$. Then the covariance matrix, $\boldsymbol{\Sigma}$, becomes $\sigma^2_v \mathbf{Z} \mathbf{Z}^\top + \sigma^2_{de}\mathbf{R} + \sigma^2_{ie}\mathbf{I}$.

Previously we compared models using \texttt{AIC()}, but another way to compare models is by leave-one-out cross validation [@hastie2009elements]. In leave-one-out cross validation, each observation is held-out and the model is re-fit and used to predict the held-out observation. Then a loss statistic is computed that compares each prediction to its true value. We calculate mean-squared-prediction error (MSPE) using leave-one-out cross validation for both the Poisson and negative binomial model by running
```{r}
loocv(count_mod_pois)
loocv(count_mod_nb)
```

The negative binomial model has a lower MSPE, which suggests that incorporating overdispersion improves model fit here. Tidying this fitted model we see that there is some evidence elevation is related to moose counts and that this effect changes across strata (all $p~$values $< 0.1$):
```{r}
tidy(count_mod_nb)
```

The \texttt{varcomp()} function in `spmodel` apportions model variability into fixed and random components:
```{r}
varcomp(count_mod_nb)
```
Most model variability is explained by the spatially dependent random error (\texttt{de}) and the random effect for strata (\texttt{1 | strat}). Running
```{r, eval = FALSE}
plot(count_mod_nb, which = c(4, 5))
```

```{r, fig.show='hold', out.width="50%"}
plot(count_mod_nb, which = 4)
plot(count_mod_nb, which = 5)
```


shows observations of high influence or leverage and that the standardized residuals tend to be spread out around zero.

Estimated mean moose counts for the \texttt{moose} data (obtained via \texttt{fitted()}) and predicted mean moose counts for the \texttt{moose\_preds} data (obtained via \texttt{predict()} or \texttt{augment()}) share similiar patterns and are overlain by running
```{r}
aug_mod <- augment(count_mod_nb)
aug_pred <- augment(count_mod_nb, newdata = moose_preds, type = "response")
aug_mod$type <- "mod"
aug_pred$type <- "pred"
keep_cols <- c(".fitted", "type")
aug_combined <- rbind(aug_mod[, keep_cols], aug_pred[, keep_cols])
ggplot(aug_combined, aes(color = .fitted, shape = type)) +
  geom_sf(size = 2) +
  scale_color_viridis_c(option = "H")  +
  theme_gray(base_size = 18)
```

# Discussion {#sec:discussion}

Throughout this vignette, we have shown several features `spmodel` offers, including a novel application of the Laplace approximation, similarity to base **R**'s \texttt{glm()} function, over a dozen spatial covariance functions, a variety of tools available to evaluate models, inspect model diagnostics, and make predictions using ubiquitous base **R** functions (e.g., \texttt{summary()}, \texttt{plot()}, and \texttt{predict()}) and more. Spatial generalized linear models for point-referenced data (i.e., generalized geostatistical models) are fit using the \texttt{spglm()} function. Spatial generalized linear models for areal data (i.e., generalized spatial autoregressive models) are fit using the \texttt{spgautor()} function. Both functions share common structure and syntax. Spatial data are simulated in `spmodel` by adding an \texttt{sp} prefix to commonly used base **R** simulation functions (e.g., \texttt{sprbinom()}). 

We appreciate feedback from users regarding `spmodel`. To learn more about how to provide feedback or contribute to `spmodel`, please visit our GitHub repository at [https://github.com/USEPA/spmodel](https://github.com/USEPA/spmodel).

# References {.unnumbered}

<div id="refs"></div>
