---
title: "Spatial Generalized Linear Models in spmodel"
author: "Michael Dumelle, Matt Higham, and Jay M. Ver Hoef"
bibliography: '`r system.file("references.bib", package="spmodel")`'
output:
  html_document:
---

```{r setup, include = FALSE}
# # jss style
# knitr::opts_chunk$set(prompt=TRUE, echo = TRUE, highlight = FALSE, continue = " + ", comment = "")
# options(replace.assign=TRUE, width=90, prompt="R> ")

# rmd style
knitr::opts_chunk$set(collapse = FALSE, comment = "#>", warning = FALSE, message = FALSE)

# loading
library(ggplot2)
library(spmodel)
```

# Introduction {#sec:introduction}

`spmodel` is an **R** package used to fit, summarize, and
predict for a variety of spatial statistical models. The vignette
provides an introduction spatial generalized linear models for
non-Gaussian response distributions in `spmodel`. Before proceeding, we
load `spmodel` by running

```{r, eval = FALSE}
library(spmodel)
```

If using `spmodel` in a formal publication or report, please cite it.
Citing `spmodel` lets us devote more resources to the package in the
future. We view the `spmodel` citation by running

```{r}
citation(package = "spmodel")
```

We will create visualizations using `ggplot2` [@wickham2016ggplot2],
which we load by running

```{r, eval = FALSE}
library(ggplot2)
```

`ggplot2` is only installed alongside `spmodel` when
`dependencies = TRUE` in `install.packages()`, so check that the package
is installed and loaded before reproducing any of these vignette's
visualizations. We will also show code that can be used to create
interactive visualizations of spatial data with `mapview`
[@appelhans2022mapview]. `mapview` also has many backgrounds available
that contextualize spatial data with topographical information. Before
running the `mapview` code interactively, make sure `mapview` is
installed and loaded.

`spmodel` contains various methods for generic functions defined outside
of `spmodel`. To find relevant documentation for these methods, run
`help("generic.spmodel", "spmodel")` (e.g.,
`help("summary.spmodel", "spmodel")`,
`help("predict.spmodel", "spmodel")`, etc.). Note that
`?generic.spmodel` is shorthand for
`help("generic.spmodel", "spmodel")`. We provide more details and
examples regarding these methods and generics throughout this vignette.
For a full list of `spmodel` functions available, see `spmodel`'s
documentation manual.

# A Review of the Spatial Linear Model

We first focus on spatial linear models for a Gaussian response vector.
The spatial linear model is defined as \begin{equation}\label{eq:splm}
 \mathbf{y} = \mathbf{X} \boldsymbol{\beta} + \boldsymbol{\tau} + \boldsymbol{\epsilon},
\end{equation} where for a sample size $n$, $\mathbf{y}$ is an
$n \times 1$ column vector of response variables, $\mathbf{X}$ is an
$n \times p$ design (model) matrix of explanatory variables,
$\boldsymbol{\beta}$ is a $p \times 1$ column vector of fixed effects
controlling the impact of $\mathbf{X}$ on $\mathbf{y}$,
$\boldsymbol{\tau}$ is an $n \times 1$ column vector of spatially
dependent random errors, and $\boldsymbol{\epsilon}$ is an $n \times 1$
column vector of spatially independent random errors. We make a few
assumptions about $\boldsymbol{\tau}$ and $\boldsymbol{\epsilon}$ in the
spatial linear model: first, that
$\text{E}(\boldsymbol{\tau}) = \text{E}(\boldsymbol{\epsilon}) = \boldsymbol{0}$,
where $\text{E}(\cdot)$ denotes expectation; second, that
$\text{Cov}(\boldsymbol{\tau}) = \sigma^2_\tau \mathbf{R}$, where
$\mathbf{R}$ is an $n \times n$ matrix that determines the spatial
dependence structure in $\mathbf{y}$ and depends on a range parameter,
$\phi$; third, that
$\text{Cov}(\boldsymbol{\epsilon}) = \sigma^2_\epsilon \mathbf{I}$,
where $\mathbf{I}$ is an $n \times n$ identity matrix; and fourth, that
$\boldsymbol{\tau}$ and $\boldsymbol{\epsilon}$ are independent of one
another. The parameter $\sigma^2_{\tau}$ is called the spatially
dependent random error variance or partial sill. The parameter
$\sigma^2_\epsilon$ is called the spatially independent random error
variance or nugget. These two variance parameters are henceforth more
intuitively written as $\sigma^2_{de}$ and $\sigma^2_{ie}$,
respectively. The covariance of $\mathbf{y}$ is denoted
$\boldsymbol{\Sigma}$ and given by \begin{equation}\label{eq:spcov}
 \boldsymbol{\Sigma} = \sigma^2_{de}\mathbf{R} + \sigma^2_{ie} \mathbf{I}.
\end{equation} The parameters $\sigma^2_{de}$, $\phi$, and
$\sigma^2_{ie}$ are elements of $\boldsymbol{\theta}$, the covariance
parameter vector.

The spatial linear model applies to both point-referenced and areal
(i.e., lattice; polygon) data. Spatial data are point-referenced when
the elements in $\mathbf{y}$ are observed at point-locations indexed by
x-coordinates and y-coordinates on a spatially continuous surface with
an infinite number of locations. For example, consider sampling soil at
any point-location in a field. Spatial linear models for
point-referenced data are sometimes called geostatistical models. An
example of an $\mathbf{R}$ matrix for point-referenced data is the
spherical correlation matrix given by
\begin{equation}\label{eq:spherical}
\mathbf{R} = \left(1 - 1.5\frac{\mathbf{H}}{\phi} + 0.5\frac{\mathbf{H}^3}{\phi^3}\right) \odot \mathcal{I}(\mathbf{H} \leq \phi),
\end{equation}\
where $\mathbf{H}$ is a matrix of Euclidean distances among
observations, $\mathbf{H}^3$ is a matrix of cubed Euclidean distances,
$\odot$ is the Hadmard (element-wise; direct) product, and
$\mathcal{I}(\mathbf{H} \leq \phi)$ is an indicator function equal to
one when the $ij$th element of $\mathbf{H}$ is less than $\phi$ (the
range) and zero otherwise. A second is example is the Matérn correlation
matrix given by \begin{equation}\label{eq:matern}
\mathbf{R} = \frac{2^{1 - \nu}}{\Gamma(\nu)} \left(2\nu \mathbf{H} / \phi \right)^{\nu / 2} \odot B_k(2\nu \mathbf{H}, \nu),
\end{equation} where $\nu$ is a smoothness parameter, $\odot$ is the
Hadmart (element-wise; direct) product, and $B_k(\cdot, \nu)$ is a
Bessel-K function with order $\nu$. @chiles2012geostatistics provide
examples of several spatial dependence structures.

Spatial data are areal when they are observed as part of a finite
network of polygons whose connections are indexed by a neighborhood
structure. For example, the polygons may represent counties in a state
who are neighbors if they share at least one boundary. An example of an
$\mathbf{R}$ matrix for areal data is the simultaneous-autoregressive
spatial correlation matrix given by \begin{equation*}
  \mathbf{R} = [(\mathbf{I} - \phi \mathbf{W})(\mathbf{I} - \phi \mathbf{W})^\top]^{-1},
\end{equation*} where $\mathbf{W}$ is a weight matrix that describes the
neighborhood structure among observations. Spatial linear models for
areal data are sometimes called spatial autoregressive models. For
thorough reviews of the spatial linear model, see
@cressie1993statistics, @banerjee2014hierarchical, and
@schabenberger2017statistical.

# The Spatial Generalized Linear Model

It is common to implicitly assume $\mathbf{y}$ in the spatial linear
model is unconstrained and Gaussian. This is a restrictive assumption
and often inappropriate for binary data, proportion data, count data,
and skewed data, which are common in many practical applications. These
types of data are most naturally described using generalized linear
models [@nelder1972generalized; @mccullagh1989generalized]. Generalized
linear models are an extension of linear models that "link" a function
of a distribution's mean to the linear function
$\mathbf{X} \boldsymbol{\beta}$ (the explanatory variables and fixed
effects). We can extend generalized linear models by adding linear,
latent (i.e., unobserved) random effects to
$\mathbf{X} \boldsymbol{\beta}$. These types of models are called
generalized linear mixed models [@breslow1993approximate] and are often
formulated using a hierarchical construction [@lee1996hierarchical]. For
thorough reviews of generalized linear models and generalized linear
mixed models, see @bolker2009generalized, @myers2012generalized,
@faraway2016extending, and @wood2017generalized.

A recent `spmodel` release (version 0.4.0) provided support for spatial
generalized linear models (i.e., a generalized linear mixed model with
spatial random effects). We define the spatial generalized linear model
as \begin{equation}\label{eq:spglm}
f(\boldsymbol{\mu}) \equiv \mathbf{w} = \mathbf{X} \boldsymbol{\beta} + \boldsymbol{\tau} + \boldsymbol{\epsilon},
\end{equation} where $f(\cdot)$ is the link function, $\boldsymbol{\mu}$
is the mean of $\mathbf{y}$, and the remaining terms $\mathbf{X}$,
$\boldsymbol{\beta}$, $\boldsymbol{\tau}$, $\boldsymbol{\epsilon}$
represent the same quantities as for spatial linear models. That is,
$\text{E}(\boldsymbol{\tau}) = \text{E}(\boldsymbol{\epsilon}) = \boldsymbol{0}$,
$\text{Cov}(\boldsymbol{\tau}) = \sigma^2_{de}\mathbf{R}$,
$\text{Cov}(\boldsymbol{\epsilon}) = \sigma^2_{ie} \mathbf{I}$,
$\boldsymbol{\tau}$ is independent of $\boldsymbol{\epsilon}$, and
$\boldsymbol{\Sigma} = \sigma^2_{de} \mathbf{R} + \sigma^2_{ie} \mathbf{I}$.
The link function, $f(\cdot)$, "links" a function of $\boldsymbol{\mu}$
to the linear term $\mathbf{w}$. Note that the linking of
$\boldsymbol{\mu}$ to $\mathbf{w}$ applies element-wise to each vector.
Each link function $f(\cdot)$ has a corresponding inverse link function,
$f^{-1}(\cdot)$, which "links" a function of $\mathbf{w}$ to
$\boldsymbol{\mu}$. Notice that $\mathbf{w}$ is unconstrained but
$\boldsymbol{\mu}$ is usually constrained in some way (e.g., positive)
that depends on the distribution assumed for $\mathbf{y}$.

Spatial linear and generalized linear models rely on the same
components: explanatory variables, fixed effects, spatially dependent
random errors, and spatial independent random errors. These components
are related to the response linearly (spatial linear models) or
non-linearly (spatial generalized linear models). The response itself is
generally assumed to follow a Gaussian distribution (spatial linear
models) or one of many non-Gaussian distributions (spatial generalized
linear models). Some non-Gaussian distributions include the binomial
distribution for binary responses, the beta distribution for proportion
responses, the Poisson or negative binomial distributions for count
responses, and the gamma or inverse Gaussian distributions for skewed,
positive continuous responses.

`spmodel` implements a marginal approach to inference that uses a novel
application of the Laplace approximation. This marginal approach is
powerful and flexible, accommodating a wide range of covariance
structures. It is a frequentist approach, which means that it does not
rely on any Bayesian sampling routines, as is common when handling
complex hierarchical structures. It also formally maximizes a
likelihood, which is advantageous because it means that likelihood-based
statistics such as AIC [@akaike1974new; @hoeting2006model] and BIC [@schwarz1978estimating] are defined,
in contrast to approaches that only specify the first two moments of a
distribution (e.g., quasi-likelihood fit via iteratively re-weighted
least squares). @ver2024marginal provide the methodology's full details.

`spmodel` accommodates the binomial, beta, Poisson, negative binomial,
gamma, and inverse Gaussian distributions. The binomial and beta
response distributions (specified by `family`) use the logit
link, while the Poisson, negative binomial, gamma, and inverse Gaussian
distributions use the log link -- see the following table.

| Family | Link Function | Link Name | Data Type |
|------|------|------|------|
| Binomial | $f(\mathbf{\mu}) = \log(\mathbf{\mu} / (1 - \mathbf{\mu}))$ | Logit | Binary; Binary Count |
| Poisson | $f(\mathbf{\mu}) = \log(\mathbf{\mu})$ | Log | Count |
| Negative Binomial | $f(\mathbf{\mu}) = \log(\mathbf{\mu})$ | Log | Count |
| Beta | $f(\mathbf{\mu}) = \log(\mathbf{\mu} / (1 - \mathbf{\mu}))$ | Logit | Proportion |
| Gamma | $f(\mathbf{\mu}) = \log(\mathbf{\mu})$ | Log | Skewed |
| Inverse Gaussian | $f(\mathbf{\mu}) = \log(\mathbf{\mu})$ | Log | Skewed | 

: Spatial generalized linear model response distributions available in
spmodel alongside their corresponding link functions and data types.

## Model fitting {#sec:model-fit}

Spatial generalized linear models in `spmodel` are fit using
`spglm()` (for point-referenced data) or `spgautor()` (for
areal data). These functions are similar in structure both to one
another and to the `glm()` function in base **R**. They generally
require the following four arguments: `formula`, a formula that
describes the relationship between the response and explanatory
variables; `family`, the response distribution, which can be
`binomial`, `beta`, `poisson`, `nbinomial`,
`Gamma`, or `inverse.gaussian`; `data`, a
`data.frame` or `sf` object [@pebesma2018sf] whose rows
index observations and whose columns contain the response variable,
explanatory variables, and spatial information; and
`spcov_type`, one of 19 spatial covariance types (e.g.,
spherical spatial covariance, Matérn spatial covariance,
simultaneous-autoregressive spatial covariance etc.). The binomial and
beta response distributions (specified by `family`) use the logit
link, while the Poisson, negative binomial, gamma, and inverse Gaussian
distributions use the log link. Other
arguments to `spglm()` and `spgautor()` are either
required only in specific situations or have default values specified.
All arguments to the `spglm()` and `spgautor()` functions
are summarized in the following table. Models for several
hundred observations can take anywhere from a few seconds to a few
minutes to fit, depending on the shape of the likelihood surface and how
quickly the covariance parameters converge. Models for several thousand
observations can utilize `spmodel`'s large data set options.

| Argument | Purposes | Applies To |
|------|------|------|
| `formula}` | Fixed effects formula | Both  | 
| `family}` | Response distribution | Both  |
| `data` | Data Frame or `sf` object | Both  |
| `spcov_type` | Spatial covariance type | Both  |
| `spcov_initial` | Initial or known spatial covariance parameters | Both  | 
| `dispersion_initial` | Initial or known dispersion parameter | Both  |
| `estmethod` | Estimation method | Both  |
| `random` | Random effects formula | Both  |
| `randcov_initial` | Initial or known random effect variances | Both  | 
| `partition_factor` | A partition factor formula | Both  |
| `...` | Additional arguments | Both |
| `anisotropy` | Whether to model anisotropy | `spglm()` |
| `xcoord` | An x-coordinate name | `spglm()}`  |
| `ycoord` | A y-coordinate name | `spglm()`  |
| `local` | Options for large data sets | `spglm()`  |
| `W` | A neighborhood weight matrix | `spgautor()`  |
| `row_st` | Whether to row-standardize `W` | `spgautor()`  |
| `M` | A symmetry matrix | `spgautor()`  |
| `range_positive` | Whether the range parameter is restricted to be positive |
| `cutoff` | Distance cutoff if using distance-based neighbor definition |

: Arguments to `spglm()` and `spgautor()`, their purpose, and whether the
argument applies to both `spglm()` and `spgautor()` or just one of them.

## Model evaluation and diagnostics {#sec:diagnostics}

After fitting a model, it is usually beneficial to evaluate certain
aspects of model fit or inspect and visualize model diagnostics like
residuals. There are many functions in `spmodel` that are used for these
purposes, and their main argument is the fitted model object returned by
`spglm()` or `spgautor()`. The names of these functions
are meant to be illustrative, describing the purpose of the function
itself. The functions for evaluating model fit
(see below) return likelihood-based
statistics like AIC, AICc, BIC, and deviance [@mccullagh1989generalized],
return coefficient estimates and confidence intervals, perform cross
validation, and more. 

| Function | Purpose |
|------|------|
| `AIC()`; `AICc()` | Return AIC and AICc   | 
| `anova()` | Perform an analysis of variance |
| `AUROC()` | Compute the area under the receiver operating characteristic curve |
| `BIC()` | Return BIC   | 
| `coef()`; `coefficients()` | Return parameter estimates   |
| `confint()` | Return confidence intervals  | 
| `covmatrix()` | Return the fitted covariance matrix   |
| `deviance()` | Return the deviance  |
| `glance()`; `glances()` | Glance at model fits | 
| `logLik()` | Return the log-likelihood  |
| `loocv()` | Perform leave-one-out cross validation |
| `pseudoR2()` | Return the pseudo R-squared   |
| `summary()` | Summarize the fitted model   |
| `tidy()` | Tidy the fitted model   |
| `varcomp()` | Compare sources of variability |
| `vcov()` | Return variance-covariance matrices  |

: Model evaluation functions available for models fit using spglm() or
spgautor() and their purpose.

The functions for inspecting model diagnostics
(see below) return Cook's distances
[@cook1979influential], fitted values, leverage (hat) values, residuals
(of several types), and visualizations.

| Function | Purpose |
|------|------|
| `augment()` | Augment model data with diagnostics |
| `cooks.distance()` | Return Cook's distances   | 
| `fitted()`; `fitted.values()` | Return fitted values   |
| `hatvalues()`; | Return hat (leverage) values |
| `plot()` | Plot model diagnostics   |
| `residuals()`; `resid()`; `rstandard()` | Return residuals  | 

: Mode diagnostic functions available for models fit using spglm() or
spgautor() and their purpose.

The functions `tidy()`, `glance()`, and
`augment()`, popularized by the `broom` **R** package
[@robinson2021broom], are particularly useful. The `tidy()`
function tidies parameter estimate output, returning a special
`data.frame` called a `tibble` [@muller2021tibble] that is
much easier to manipulate and work with than the parameter estimate
output printed by `summary()` to the **R** console. The
`glance()` function glances at the model fit, returning a
`tibble` with the sample size, number of estimated parameters,
and several model-fit statistics. `glances()` is an extension of
`glance()` that glances at multiple models simultaneously. The
`augment()` function augments the model fit, returning a
`tibble` with the data used to fit the model as well as model
diagnostics. The `augment()` function can also be used to augment
prediction data, as we discuss next.

## Prediction {#sec:prediction}

Often a priority of a spatial data analysis is using the fitted model to
make predictions at unobserved locations and quantify uncertainties in
those predictions. The `predict()` function in `spmodel` is used
to predict the mean process at unobserved locations. It is similar in
structure to `predict()` for objects fit using `glm()` in
base **R** and takes two main arguments: `object`, the fitted
model object returned by `spglm()` or `spgautor()`; and
`newdata`, a `data.frame` or `sf` object whose rows
index locations requiring prediction and whose columns contain the
explanatory variables and spatial information at these locations. Other
arguments to `predict()` are either required only in specific
situations or have default values specified. All arguments to the
`predict()` function are summarized in
the following table. The `augment()` function can also
be used to augment `newdata` with predictions, taking the same
arguments as `predict()`.

| Argument | Purpose |
|------|------|
| `object` | Fitted model object |
| `newdata` | New data frame or `sf` object   |
| `type` | Whether predictions are on the link or response scale |
| `se.fit` | Whether standard errors should be returned   |
| `interval` | Interval type (prediction or confidence) |
| `level` | Prediction or confidence level   |
| `local` | Options for large data sets  | 
| `var_correct` | Whether prediction variances should be adjusted |
| `...` | Additional arguments |

: Arguments to `predict()` and their purpose.

## Simulation {#sec:simulation}

`spmodel` has functions that are used
to simulate spatial data from Gaussian, binomial, beta, Poisson,
negative binomial, gamma, and inverse Gaussian response distributions.
These simulation functions are similar in structure to commonly used
base **R** functions for simulating data. They are also similar in name,
simply adding an `sp` prefix to the conventional name in base
**R**. For example, `rpois()` is used to simulate Poisson data in
base **R**, while `sprpois()` is used to simulate spatial Poisson
data in `spmodel`. These functions are summarized in the following table.

| Function | Purpose |
|------|------|
| `sprnorm()` | Simulate spatial Gaussian data |
| `sprbinom()` | Simulate spatial binomial data |
| `sprbeta()` | Simulate spatial beta data |
| `sprpois()` | Simulate spatial Poisson data |
| `sprnbinom()` | Simulate spatial negative binomial data |
| `sprgamma()` | Simulate spatial gamma data |
| `sprinvgauss()` | Simulate spatial inverse Gaussian data |

: Functions to simulate spatial data and their purpose.

# An application to binary (presence/absence) data {#sec:application-binary}

In the next two sections we to show how to use `spmodel` to analyze data
collected on moose in the Togiak region of Alaska, USA. The
`moose` data in `spmodel` contain information on moose counts
from a survey of sites in the Togiak region. The data are stored as an
`sf` object, a special `data.frame` built to store spatial data.
`moose` has several variables: `strat`; a stratification
variable with two levels (`L` for low and `M` for medium)
based on surrounding landscape metrics; `elev`, the site
elevation; `presence`, whether or not at least one moose was
observed at the site; `count`, the number of moose observed at a
the site; and `geometry`, the spatial coordinates in an Alaska
Albers projection (EPSG: 3338). We load `spmodel` and the `moose`
data by running

```{r}
library("spmodel")
data("moose")
```

We are interested in quantifying the effects of `elev` and
`strat` on moose `presence` and begin with some
visualizations. First, we visualize the presence for observed sites:

```{r moosepres, out.width="65%", fig.align="center", fig.cap="Moose presence."}
ggplot(moose, aes(color = presence)) + 
  geom_sf(size = 2) + 
  scale_color_viridis_d(option = "H") +
  theme_gray(base_size = 14)
```

There is spatial patterning in the distribution of moose presence, as
moose tend to be present in the eastern and southwestern parts of the
domain. This spatial pattern seems to have a directional orientation,
seemingly strongest in the northwest-to-southeast direction.

Next, we visualize the elevation for each strata:

```{r moosestrat, out.width="100%", fig.cap="Elevation versus strata in the moose data."}
ggplot(moose, aes(color = elev)) +
  geom_sf(size = 1) + 
  facet_grid(~ strat) +
  scale_color_viridis_c()  +
  scale_y_continuous(breaks = seq(62.6, 63.6, length.out = 5)) +
  scale_x_continuous(breaks = seq(-148.5, -146, length.out = 3)) +
  theme_gray(base_size = 12)
```

There are no obvious spatial patterns in the assignment of sites to
strata, while elevation is highest on the boundaries of the domain and
lowest in the middle of the domain. There is spatial patterning in the
distribution of moose presence, as moose tend to be present in the
eastern and southwestern parts of the domain. This spatial pattern seems
to have a directional orientation, seemingly strongest in the
northwest-to-southeast direction.

To quantify the effects of `elev`, `strat`, and spatial
dependence on moose presence, we fit three binomial (i.e., ) logistic
regression models using `spglm()`:

```{r}
# model one
bin_mod <- spglm(
  formula = presence ~ elev + strat + elev:strat,
  family = "binomial",
  data = moose,
  spcov_type = "none"
)
# model two
bin_spmod <- spglm(
  formula = presence ~ elev + strat + elev:strat,
  family = "binomial",
  data = moose,
  spcov_type = "spherical"
)
# model three
bin_spmod_anis <- spglm(
  formula = presence ~ elev + strat + elev:strat,
  family = "binomial",
  data = moose,
  spcov_type = "spherical",
  anisotropy = TRUE
)
```

All three models have the same fixed effect structure, using elevation,
strata, and their interaction to explain moose presence. The three
models vary in their covariance structure. The first model,
`bin_mod`, has no spatial covariance ($\sigma^2_{de} = 0$). The
second model, `bin_spmod` has a spherical spatial covariance.
The third model, `bin_spmod_anis`, has a spherical spatial
covariance that incorporates geometric anisotropy. Geometric anisotropy
allows the spatial covariance to vary with direction by evaluating the
spatial covariance with $\mathbf{H}^*$ (instead of $\mathbf{H}$), a
matrix of distances between transformed coordinates that are rotated and
scaled appropriately [@schabenberger2017statistical]. We evaluate the
three models using AIC by running

```{r}
AIC(bin_mod, bin_spmod, bin_spmod_anis)
```

The spatial models (`bin_spmod`, `bin_spmod_anis`) have
a much lower AIC than the non-spatial model (`bin_mod`), which
suggests that the models benefit from incorporating spatial dependence.
`bin_spmod_anis` has a lower AIC than `bin_spmod`,
which suggests that the model benefits from incorporating directionality
in the spatial dependence. Next we inspect `bin_spmod_anis` and
later use it to predict moose presence probability at unobserved sites.

We summarize `bin_spmod_anis` using `summary()`:
```{r}
summary(bin_spmod_anis)
```

The `summary()` output from an `spglm()` model is very
similar to the `summary()` output from a base-**R**
`glm()` model, returning the original function call, deviance
residuals, a fixed effects coefficients table, a pseudo R-squared (which
quantifies the amount of variability explained by the fixed effects),
and covariance parameter coefficients. The fixed effects coefficients
table provides some evidence that elevation and strata are associated
with with moose presence and that the effect of elevation varies across
the two strata (all $p~$values $<$ 0.05). 

The area under the receiver receiver operating characteristic (AUROC) curve quantifies the effectiveness of a classifier [@robin2011proc]. It ranges from zero to one, and higher values indicate better model performance. The AUROC of `bin_spmod_anis` is
```{r}
AUROC(bin_spmod_anis)
```

The `plot()` function
returns graphics of several model diagnostics. Running

```{r, eval = FALSE}
plot(bin_spmod_anis, which = c(7, 8))
```

```{r, echo = FALSE, fig.show='hold', out.width="50%", fig.cap="Spatial dependence is a function of distance (left) and as an anisotropic level curve of equal correlation (right)."}
plot(bin_spmod_anis, which = 7)
plot(bin_spmod_anis, which = 8)
```

shows that the spatial dependence is evident and appears strongest in
the northwest-to-southeast direction. These findings are supported by
the clear spatial patterns in the data.

Recall that the `tidy()`, `glance()`, and
`augment() functions are particularly useful tools for
manipulating and understanding fitted model objects. The `tidy()`
function tidies model output, returning a `tibble` of parameter
estimates (and confidence intervals):

```{r}
tidy(bin_spmod_anis, conf.int = TRUE)
```

By default, the fixed effects estimates are returned, but covariance
parameter estimates are returned via the `effects` argument:

```{r}
tidy(bin_spmod_anis, effects = "spcov")
```

The `is_known` column indicates whether covariance parameters
were assumed known, which is possible to specify using the
`spcov_initial` argument to `spglm()`. Here, all
parameters were assumed unknown and then estimated.

The `glance()` function glances at model fit, returning a
`tibble` of model fit statistics:

```{r}
glance(bin_spmod_anis)
```

`glances()` can be used to glance at multiple models
simultaneously and sorts models by ascending AICc:

```{r}
glances(bin_mod, bin_spmod, bin_spmod_anis)
```

The `augment()` function augments the data with model
diagnostics:

```{r}
aug_mod <- augment(bin_spmod_anis)
aug_mod
```

where `.fitted` are fitted values (the estimated
$f^{-1}(\mathbf{w})$, or moose presence probability), `.resid`
are response residuals, `.hat` values indicate leverage,
`.cooksd` values indicate Cook's distance (i.e., influence), and
`.std.resid` are standardized residuals.

The `moose_preds` data contains survey sites that were not
sampled. We can use `bin_spmod_anis` to make predictions of the
underlying probabilities of moose presence at these sites using
`predict()` or `augment()`. `predict()` and
`augment()` return the same predictions but `augment()`
augments the prediction data with the predictions:

```{r}
data("moose_preds")

spmod_anis_preds <- predict(
  bin_spmod_anis,
  newdata = moose_preds,
  type = "response",
  interval = "prediction"
)
head(spmod_anis_preds)

aug_pred <- augment(
  bin_spmod_anis,
  newdata = moose_preds,
  type.predict = "response", 
  interval = "prediction"
)
aug_pred
```

Estimated moose presence probability for the `moose` data
(obtained via `fitted()`) and predicted moose presence
probability for the `moose_preds` data (obtained via
`predict()` or `augment()`) are overlain below and share
similar patterns:

```{r bin, out.width="65%", fig.align="center", fig.cap="Fitted values and predictions for moose presence probability."}
aug_mod$type <- "mod"
aug_pred$type <- "pred"
keep_cols <- c(".fitted", "type")
aug_combined <- rbind(aug_mod[, keep_cols], aug_pred[, keep_cols])
ggplot(aug_combined, aes(color = .fitted, shape = type)) +
  geom_sf(size = 2) +
  scale_color_viridis_c(option = "H")  +
  theme_gray(base_size = 14)
```

# An application to count data {#sec:application-count}

The `moose` data also contain the number of moose observed at
each site:

```{r moose, out.width="65%", fig.align="center", fig.cap="Moose counts."}
ggplot(moose, aes(color = count)) +
  geom_sf(size = 2) +
  scale_color_viridis_c(option = "H")  +
  theme_gray(base_size = 14)
```

The moose counts are similarly distributed as moose presence, highest in
the eastern and southwestern parts of the domain. We compare two count
models, Poisson and negative binomial. The Poisson model assumes the
underlying process generating the counts has the same mean and variance
while the negative binomial model allows for overdispersion (i.e., the
variance is greater than the mean). We are interested in quantifying the
effects of `elev` and `strat` on moose `count`, albeit
using a slightly different approach than we did for moose `presence`.
Here, we will model elevation as a fixed effect and allow this effect to
change between strata, but we will model strata as a random effect to
highlight additional flexibility of the `spmodel` package. We fit
relevant Poisson and negative binomial models with a Matérn spatial
covariance using `spglm()`:

```{r}
count_mod_pois <- spglm(
  count ~ elev + elev:strat,
  family = "poisson",
  data = moose,
  spcov_type = "matern",
  random = ~ (1 | strat)
)

count_mod_nb <- spglm(
  count ~ elev + elev:strat,
  family = "nbinomial",
  data = moose,
  spcov_type = "matern",
  random = ~ (1 | strat)
)
```

Random effects are specified in `spmodel` via the `random`
argument using similar syntax as the commonly used `lme4`
[@bates2015lme4] and `nlme` [@pinheiro2006mixed] **R** packages for
non-spatial mixed models. In `spmodel`, `~ strat` is short-hand
for `~ (1 | strat)`. Random effects alter the covariance
structure of the model, building additional correlation into the model
for sites sharing a level of the random effect (here, sites within the
same strata). More formally, when incorporating a random effect, the
spatial generalized linear model becomes
\begin{equation}\label{eq:spglm-rand}
f(\boldsymbol{\mu}) \equiv \mathbf{w} = \mathbf{X} \boldsymbol{\beta} + \mathbf{Z}\mathbf{v} + \boldsymbol{\tau} + \boldsymbol{\epsilon},
\end{equation} where $\mathbf{Z}$ is a design matrix that indexes the
random effects, $\mathbf{v}$, and
$\text{Cov}(\mathbf{v}) = \sigma^2_{v}\mathbf{I}$. Then the covariance
matrix, $\boldsymbol{\Sigma}$, becomes
$\sigma^2_v \mathbf{Z} \mathbf{Z}^\top + \sigma^2_{de}\mathbf{R} + \sigma^2_{ie}\mathbf{I}$.

Previously we compared models using `AIC()`, but another way to
compare models is by leave-one-out cross validation
[@hastie2009elements]. In leave-one-out cross validation, each
observation is held-out and the model is re-fit and used to predict the
held-out observation. Then a loss statistic is computed that compares
each prediction to its true value. We calculate mean-squared-prediction
error (MSPE) using leave-one-out cross validation for both the Poisson
and negative binomial model by running

```{r}
loocv(count_mod_pois)
loocv(count_mod_nb)
```

The negative binomial model has a lower MSPE, which suggests that
incorporating overdispersion improves model fit here. Tidying this
fitted model we see that there is some evidence elevation is related to
moose counts and that this effect changes across strata (all $p~$values
$< 0.1$):

```{r}
tidy(count_mod_nb)
```

The `varcomp()` function in `spmodel` apportions model
variability into fixed and random components:

```{r}
varcomp(count_mod_nb)
```

Most model variability is explained by the spatially dependent random
error (`de`) and the random effect for strata
(`1 | strat`). Running

```{r, eval = FALSE}
plot(count_mod_nb, which = c(4, 5))
```

```{r, echo = FALSE, fig.show='hold', out.width="50%", fig.cap="Model diagnostics."}
plot(count_mod_nb, which = 4)
plot(count_mod_nb, which = 5)
```

shows observations of high influence or leverage and that the
standardized residuals tend to be spread out around zero.

Estimated mean moose counts for the `moose` data (obtained via
`fitted()`) and predicted mean moose counts for the
`moose_preds` data (obtained via `predict()` or
`augment()`) share similar patterns and are overlain by running

```{r nb, out.width="65%", fig.align="center", fig.cap="Fitted values and predictions for moose counts."}
aug_mod <- augment(count_mod_nb)
aug_pred <- augment(count_mod_nb, newdata = moose_preds, type.predict = "response")
aug_mod$type <- "mod"
aug_pred$type <- "pred"
keep_cols <- c(".fitted", "type")
aug_combined <- rbind(aug_mod[, keep_cols], aug_pred[, keep_cols])
ggplot(aug_combined, aes(color = .fitted, shape = type)) +
  geom_sf(size = 2) +
  scale_color_viridis_c(option = "H")  +
  theme_gray(base_size = 18)
```

# Discussion {#sec:discussion}

Throughout this vignette, we have shown several features `spmodel`
offers, including a novel application of the Laplace approximation,
similarity to base **R**'s `glm()` function, over a dozen spatial
covariance functions, a variety of tools available to evaluate models,
inspect model diagnostics, and make predictions using ubiquitous base
**R** functions (e.g., `summary()`, `plot()`, and
`predict()`) and more. Spatial generalized linear models for
point-referenced data (i.e., generalized geostatistical models) are fit
using the `spglm()` function. Spatial generalized linear models
for areal data (i.e., generalized spatial autoregressive models) are fit
using the `spgautor()` function. Both functions share common
structure and syntax. Spatial data are simulated in `spmodel` by adding
an `sp` prefix to commonly used base **R** simulation functions
(e.g., `sprbinom()`).

We appreciate feedback from users regarding `spmodel`. To learn more
about how to provide feedback or contribute to `spmodel`, please visit
our GitHub repository at <https://github.com/USEPA/spmodel>.

# References {.unnumbered}

::: {#refs}
:::
